---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# `{rmb}` datasets for *Regression Methods in Biostatistics*

<!-- badges: start -->

```{r, echo = FALSE, message = FALSE, results='asis'}
library(badger)
cat(
  badge_lifecycle(stage = "stable", alt = "Lifecycle: stable"),
  badge_github_actions(alt = "R-CMD-check", action = "R-CMD-check.yaml"),
  badge_codecov(alt = "Codecov test coverage", branch = "main"),
  badge_codefactor(alt = "CodeFactor", ref = "d-morrison/rmb"), # case matters
  badge_cran_release(alt = "CRAN status"),
  badge_cran_download(type = "grand-total"),
  badge_cran_download(type = "last-month"),
  badge_cran_download(type = "last-week"),
  badger::badge_license()
)
```

<!-- badges: end -->

`{rmb}` provides the 26 datasets from the companion website for
[*Regression Methods in Biostatistics* (2nd edition)](https://www.routledge.com/Regression-Methods-in-Biostatistics-Linear-Logistic-Survival-and-Repeated-Measures-Models/Vittinghoff-Glidden-Shiboski-McCulloch/p/book/9781466504401)
by Vittinghoff, Glidden, Shiboski, and McCulloch, along with full variable
labels, documentation, and example analyses organized as mini-journal articles.

## Installation

You can install the released version of `{rmb}` from [CRAN](https://cran.r-project.org/) with:

``` r
install.packages("rmb")
```

Or install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("d-morrison/rmb")
```

## Example

```{r example}
library(rmb)
head(hers[, c("HT", "age", "BMI", "glucose")])
```

## Datasets

```{r datasets, echo = FALSE}
rmb_datasets[, c("object", "title", "chapter")]
```

## Development

### Building the Documentation Site

This package uses [altdoc](https://altdoc.etiennebacher.com/) with [Quarto](https://quarto.org/) to build its documentation site. To build and preview the documentation locally:

```r
pkgload::load_all()
altdoc::render_docs()
altdoc::preview_docs()
```

The documentation is automatically built and deployed to GitHub Pages via GitHub Actions when changes are pushed to the main branch.

## Code of Conduct

Please note that the `{rmb}` project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
