---
title: "WCGS data"
format: html
vignette: >
  %\VignetteIndexEntry{WCGS data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmb)
```

This article reproduces key WCGS analyses from RMB2e Chapters 2, 3, and 5.

## Study design

```{r study-design}
data(rmb_datasets, package = "rmb")
rmb_datasets$study_design[rmb_datasets$object == "wcgs"]
```

```{r load-wcgs}
data(wcgs, package = "rmb")
dat <- wcgs
dim(dat)
summary(dat[c("sbp", "weight", "chd69", "age", "chol", "bmi", "smoke", "behpat")])
```

## RMB2e analysis workflow

### Research question

How are established risk factors (age, cholesterol, systolic blood pressure, BMI, smoking, and behavior pattern) associated with CHD risk in WCGS?

RMB2e references: SBP distribution and two-variable EDA (pp. 30–32, 39, 41), CHD table summaries (pp. 64, 69), logistic CHD modeling with risk factors and behavior pattern (pp. 171, 175–177).

### Causal assumptions (conceptual DAG)

```{mermaid}
flowchart LR
  AGE[Age] --> CHD[CHD by follow-up]
  CHOL[Cholesterol] --> CHD
  SBP[Systolic BP] --> CHD
  BMI[BMI] --> CHD
  SMOKE[Smoking] --> CHD
  BEH[Behavior pattern] --> SMOKE
  BEH --> CHD
  AGE --> SBP
  AGE --> CHOL
```

### Step 0: Exploratory data analysis

RMB2e emphasizes one- and two-variable summaries before modeling (pp. 30–41).

```{r exploratory-summaries}
summary(dat$sbp)
summary(dat$weight)
with(dat, cor(sbp, weight, use = "complete.obs"))
with(dat, table(behpat, useNA = "ifany"))
with(dat, prop.table(table(chd69)))
```

```{r exploratory-plots, fig.height=4, fig.width=6}
hist(dat$sbp, breaks = 15, col = "grey85", border = "white", main = "WCGS systolic blood pressure", xlab = "SBP")
plot(dat$weight, dat$sbp, pch = 16, cex = 0.5, col = rgb(0, 0, 0, 0.2),
     xlab = "Weight", ylab = "SBP", main = "SBP vs weight")
lines(stats::lowess(dat$weight, dat$sbp, f = 0.25), col = "#d95f02", lwd = 2)
```

### Step 1: Model specification

Following RMB2e Chapter 5, we fit logistic regression for CHD with transformed continuous predictors and categorical behavior pattern (pp. 171, 175).

```{r model-specification}
ana <- subset(dat, chol < 645)
ana$age_10 <- ana$age / 10
ana$chol_50 <- ana$chol / 50
ana$sbp_50 <- ana$sbp / 50
ana$bmi_10 <- ana$bmi / 10
ana$smoke <- factor(ana$smoke)
ana$behpat <- factor(ana$behpat)

formula_base <- chd69 ~ age_10 + chol_50 + sbp_50 + bmi_10 + smoke
formula_behpat <- update(formula_base, . ~ . + behpat)

list(base = formula_base, with_behpat = formula_behpat)
```

### Step 2: Parameter estimation

```{r parameter-estimation}
fit_base <- stats::glm(formula_base, data = ana, family = stats::binomial())
fit_behpat <- stats::glm(formula_behpat, data = ana, family = stats::binomial())

summary(fit_base)
summary(fit_behpat)
```

### Step 3: Model adequacy checks

```{r adequacy-checks}
# Nested-model comparison for contribution of behavior pattern (RMB2e pp. 175-177)
stats::anova(fit_base, fit_behpat, test = "Chisq")

# Simple calibration check via grouped observed vs fitted CHD risk
pred <- stats::fitted(fit_behpat)
cut_pred <- cut(pred, breaks = stats::quantile(pred, probs = seq(0, 1, by = 0.1)), include.lowest = TRUE)
obs <- tapply(ana$chd69, cut_pred, mean)
exp <- tapply(pred, cut_pred, mean)
data.frame(decile = names(obs), observed = as.numeric(obs), expected = as.numeric(exp))
```

### Step 4: Inference

```{r inference}
or <- exp(stats::coef(fit_behpat))
ci <- exp(stats::confint(fit_behpat))
inf <- data.frame(
  term = names(or),
  odds_ratio = unname(or),
  conf_low = ci[, 1],
  conf_high = ci[, 2],
  p_value = summary(fit_behpat)$coefficients[, "Pr(>|z|)"]
)
inf
```

### Scientific conclusions

Consistent with RMB2e Chapter 5 (pp. 171, 175–177), older age, higher cholesterol, higher systolic blood pressure, higher BMI, and smoking are associated with increased CHD odds in WCGS. Behavior pattern contributes additional information about CHD risk beyond these established predictors.
