---
title: "SOF data"
format: html
vignette: >
  %\VignetteIndexEntry{SOF data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmb)
```

This article applies a cross-sectional linear regression to the SOF dataset to examine whether BMI is associated with total bone mineral density after adjustment for age and menopausal age in older women (RMB2e Chapter 7).

## Introduction

The Study of Osteoporotic Fractures (SOF) enrolled over 9,000 community-dwelling women
aged 65 and older at four US clinical centers beginning in 1986 (Cummings et al. 1995).
Bone mineral density (BMD) declines with age after menopause;
however, body mass index (BMI) is protective for BMD
because adipose tissue produces estrogen
and because mechanical loading from higher body weight stimulates bone formation.
Disentangling the age, menopausal history, and BMI contributions to BMD
requires multipredictor regression (RMB2e Ch. 7).

```{r study-design}
data(rmb_datasets, package = "rmb")
rmb_datasets$study_design[rmb_datasets$object == "sof"]
```

Is higher BMI associated with greater total bone mineral density
after adjustment for age and menopausal age in older women?

### Causal assumptions

```{r dag, fig.height=4, fig.width=6}
set.seed(42)
dag <- ggdag::dagify(
  totbmd ~ bmi + age + meno_age,
  bmi ~ age,
  labels = c(
    totbmd = "Total BMD",
    bmi = "BMI",
    age = "Age",
    meno_age = "Menopausal age"
  ),
  exposure = "bmi",
  outcome = "totbmd"
)
ggdag::ggdag(dag, use_labels = "label", text = FALSE) +
  ggdag::theme_dag_blank() +
  ggplot2::labs(title = "SOF: Causal DAG")
```

## Methods

### Study sample

```{r load-data}
data(sof, package = "rmb")
dat <- haven::zap_labels(sof)
dim(dat)
baseline_sof <- subset(dat, visit == 2)
dim(baseline_sof)
summary(baseline_sof[c("totbmd", "bmi", "age", "meno_age")])
```

### Statistical analysis

A linear regression model is fitted at the baseline visit (visit 2)
to estimate the cross-sectional association between BMI and total BMD,
adjusting for age and menopausal age.
Using only baseline visits avoids correlation from repeated measurements (RMB2e Ch. 7).

```{r model-specification}
formula_main <- totbmd ~ bmi + age + meno_age
formula_main
```

## Results

### Descriptive statistics

```{r exploratory-summaries}
summary(baseline_sof[c("totbmd", "bmi", "age", "meno_age")])
with(baseline_sof, cor(cbind(totbmd, bmi, age, meno_age), use = "complete.obs"))
```

```{r exploratory-plot, fig.height=4, fig.width=6}
ggplot2::ggplot(baseline_sof, ggplot2::aes(x = bmi, y = totbmd)) +
  ggplot2::geom_point(alpha = 0.2, size = 0.8) +
  ggplot2::geom_smooth(method = "lm", se = FALSE, color = "#d95f02", linewidth = 1) +
  ggplot2::labs(
    title = "SOF: Total BMD vs BMI at baseline",
    x = expression("BMI (kg/m"^2*")"),
    y = "Total BMD (g/cm²)"
  ) +
  ggplot2::theme_minimal()
```

### Model estimates

```{r parameter-estimation}
fit <- stats::lm(formula_main, data = baseline_sof)
summary(fit)
```

### Model diagnostics

```{r adequacy-checks, fig.height=4, fig.width=6}
fit_data <- data.frame(
  fitted = stats::fitted(fit),
  residuals = stats::residuals(fit),
  std_residuals = stats::rstandard(fit)
)

ggplot2::ggplot(fit_data, ggplot2::aes(x = fitted, y = residuals)) +
  ggplot2::geom_point() +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  ggplot2::geom_smooth(se = FALSE, color = "blue") +
  ggplot2::labs(
    title = "Residuals vs Fitted",
    x = "Fitted values",
    y = "Residuals"
  ) +
  ggplot2::theme_minimal()

ggplot2::ggplot(fit_data, ggplot2::aes(sample = std_residuals)) +
  ggplot2::stat_qq() +
  ggplot2::stat_qq_line(color = "red") +
  ggplot2::labs(
    title = "Normal Q-Q",
    x = "Theoretical Quantiles",
    y = "Standardized residuals"
  ) +
  ggplot2::theme_minimal()
```

### Inference

```{r inference}
ci <- stats::confint(fit)
coefs <- summary(fit)$coefficients
data.frame(
  term = rownames(coefs),
  estimate = coefs[, "Estimate"],
  conf_low = ci[, 1],
  conf_high = ci[, 2],
  p_value = coefs[, "Pr(>|t|)"]
)
```

## Discussion

Higher BMI is positively associated with total BMD at baseline
after adjustment for age and menopausal age in SOF women,
consistent with the mechanical loading and estrogenic pathways described above (RMB2e Ch. 7).
Age is negatively associated with BMD, reflecting postmenopausal bone loss.
Menopausal age contributes independently,
as women who underwent menopause later experienced fewer years of estrogen deficiency.
This cross-sectional analysis uses only Visit 2 data
(the earliest available visit in the SOF dataset)
to avoid the need for mixed-model methods
introduced for the longitudinal SOF analyses elsewhere in RMB2e.

## Source

- Cummings SR et al. (1995). Risk factors for hip fracture in white women. *NEJM*, 332(12), 767–773.
- UCSF Regression Methods companion data: https://regression.ucsf.edu/sites/g/files/tkssra16191/files/wysiwyg/home/data/sof.dta
- Book: Vittinghoff E, Glidden DV, Shiboski SC, McCulloch CE (2012). *Regression Methods in Biostatistics* (2nd edition).
