---
title: "Fracture risk data"
format: html
vignette: >
  %\VignetteIndexEntry{Fracture risk data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmb)
```

This article uses the FIT GLM dataset to test whether the treatment benefit of alendronate on fracture risk differs between women with and without high fall risk, illustrating effect modification in logistic regression (RMB2e Chapter 8).

## Introduction

Alendronate reduces fracture risk by increasing BMD and improving bone microarchitecture;
however, fracture risk is also determined by the propensity to fall,
which is independent of bone strength.
Women at high fall risk may experience less relative benefit from BMD-improving treatments
because many of their fractures would occur even with stronger bones.
Testing for a treatment-by-fall-risk interaction
addresses whether the alendronate benefit is heterogeneous across fall risk categories,
a key question for precision prevention (RMB2e Ch. 8 / Black 1996).

```{r study-design}
data(rmb_datasets, package = "rmb")
rmb_datasets$study_design[rmb_datasets$object == "fitglm"]
```

Does the treatment benefit of alendronate on fracture risk differ
between women with and without high fall risk?

### Causal assumptions

```{mermaid}
flowchart LR
  TRT[Alendronate] --> FRX[Fracture]
  FALLRISK[Fall risk] --> FRX
  AGE --> FRX
  BMD[Hip BMD] --> FRX
  TRT --> INT[TRT x fall risk]
  FALLRISK --> INT
  INT --> FRX
```

## Methods

### Study sample

```{r load-data}
data(fitglm, package = "rmb")
dat <- fitglm
dim(dat)
summary(dat[c("trt01", "frx", "riskcat4", "ra_age", "htotbmd")])
```

### Statistical analysis

Logistic regression with a treatment-by-fall-risk interaction term
tests whether the alendronate effect on any fracture differs by fall risk category,
adjusted for age and total hip BMD (RMB2e Ch. 8).
The interaction term directly encodes effect modification on the log-odds scale.

```{r model-specification}
formula_main <- frx ~ trt01 * riskcat4 + ra_age + htotbmd
formula_main
```

## Results

### Descriptive statistics

```{r exploratory-summaries}
with(dat, table(trt01, frx))
with(dat, prop.table(table(trt01, frx), margin = 1))
with(dat, table(riskcat4))
with(dat, tapply(htotbmd, riskcat4, mean, na.rm = TRUE))
```

```{r exploratory-plot, fig.height=4, fig.width=6}
frac_rates <- with(dat, tapply(frx, list(riskcat4, trt01), mean, na.rm = TRUE))
barplot(t(frac_rates),
        beside = TRUE,
        col = c("#1b9e77", "#d95f02"),
        xlab = "Fall risk category",
        ylab = "Fracture proportion",
        main = "FIT: Fracture rate by fall risk and treatment",
        legend.text = c("Placebo", "Alendronate"))
```

### Model estimates

```{r parameter-estimation}
fit_model <- stats::glm(formula_main, data = dat, family = stats::binomial())
summary(fit_model)
```

### Model diagnostics

```{r adequacy-checks}
fit_no_int <- stats::glm(frx ~ trt01 + riskcat4 + ra_age + htotbmd,
                         data = dat, family = stats::binomial())
stats::anova(fit_no_int, fit_model, test = "Chisq")
```

### Inference

```{r inference}
or <- exp(stats::coef(fit_model))
ci <- exp(stats::confint(fit_model))
data.frame(
  term = names(or),
  odds_ratio = unname(or),
  conf_low = ci[, 1],
  conf_high = ci[, 2],
  p_value = summary(fit_model)$coefficients[, "Pr(>|z|)"]
)
```

## Discussion

The interaction between treatment and fall risk category tests whether
the fracture-prevention benefit of alendronate is uniform or varies with fall risk.
The likelihood-ratio test for the interaction provides evidence about effect modification
on the log-odds scale, as described in RMB2e Chapter 8.
If the interaction is statistically meaningful,
subgroup-specific treatment recommendations may be appropriate;
if not, a single pooled estimate describes the treatment effect adequately.
Lower hip BMD and older age are independently predictive of fracture
regardless of treatment assignment.

## Source

- Black DM et al. (1996). Randomised trial of effect of alendronate on risk of fracture in women with existing vertebral fractures. *Lancet*, 348(9041), 1535–1541.
- UCSF Regression Methods companion data: https://regression.ucsf.edu/sites/g/files/tkssra16191/files/wysiwyg/home/data/fitglm.dta
- Book: Vittinghoff E, Glidden DV, Shiboski SC, McCulloch CE (2012). *Regression Methods in Biostatistics* (2nd edition).
