---
title: "Fecal fat data"
format: html
vignette: >
  %\VignetteIndexEntry{Fecal fat data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmb)
```

This article analyzes the fecal fat crossover experiment, testing whether different pill types affect mean fecal fat output while accounting for within-subject correlation (RMB2e Chapter 7).

## Introduction

Fecal fat excretion is a sensitive marker of intestinal fat malabsorption
and has been used to evaluate the efficacy of pancreatic enzyme replacement
and other GI interventions.
This small crossover experiment assigned subjects to different pill formulations sequentially,
measuring daily fecal fat (g/day) under each condition.
The repeated-measures design means that fecal fat values from the same subject are correlated;
failure to account for this within-subject variability
would inflate standard errors or deflate them inappropriately (RMB2e Ch. 7).

```{r study-design}
data(rmb_datasets, package = "rmb")
rmb_datasets$study_design[rmb_datasets$object == "fecfat"]
```

Do different pill types differ in mean fecal fat output, accounting for within-subject correlation?

### Causal assumptions

```{r dag, fig.height=4, fig.width=6}
set.seed(42)
dag <- ggdag::dagify(
  fecfat ~ pill + subject,
  labels = c(
    fecfat = "Fecal fat (g/day)",
    pill = "Pill type",
    subject = "Subject/patient"
  ),
  exposure = "pill",
  outcome = "fecfat"
)
ggdag::ggdag(dag, use_labels = "label", text = FALSE) +
  ggdag::theme_dag_blank() +
  ggplot2::labs(title = "Fecal fat: Causal DAG")
```

## Methods

### Study sample

```{r load-data}
data(fecfat, package = "rmb")
dat <- fecfat
dim(dat)
summary(haven::zap_labels(dat[c("fecfat", "pilltype", "subject")]))
```

### Statistical analysis

A linear model with `subject` included as a blocking factor is used
to remove between-subject variability and isolate the pill type effect,
equivalent to a one-way ANOVA within subjects (RMB2e Ch. 7).

```{r model-specification}
dat$subject <- factor(dat$subject)
dat$pilltype <- factor(dat$pilltype)
formula_main <- fecfat ~ pilltype + subject
formula_main
```

## Results

### Descriptive statistics

```{r exploratory-summaries}
with(dat, tapply(fecfat, pilltype, summary))
with(dat, tapply(fecfat, subject, mean))
```

```{r exploratory-plot, fig.height=4, fig.width=6}
ggplot2::ggplot(dat, ggplot2::aes(x = pilltype, y = fecfat)) +
  ggplot2::geom_boxplot(fill = "#d95f02") +
  ggplot2::labs(
    title = "Fecal fat by pill type",
    x = "Pill type",
    y = "Fecal fat (g/day)"
  ) +
  ggplot2::theme_minimal()
```

The swimmer plot below shows the treatment sequence for each of the 6 subjects,
with colored symbols indicating which pill type was assigned
in each of the four periods.

```{r swimplot, fig.height=4, fig.width=6}
dat_swim <- as.data.frame(dat)
dat_swim$pilltype_int <- as.integer(unclass(dat_swim$pilltype))
dat_swim$Pill_type <- factor(
  dat_swim$pilltype_int,
  levels = 1:4,
  labels = c("None", "Tablet", "Capsule", "Coated")
)
dat_swim$subject_int <- as.integer(dat_swim$subject)
dat_swim$period <- as.integer(
  ave(seq_len(nrow(dat_swim)), dat_swim$subject_int, FUN = seq_along)
)

# One row per subject for the background bar
dat_sum <- data.frame(
  subject_int = sort(unique(dat_swim$subject_int)),
  end = 4.5
)

swimplot::swimmer_plot(
  df = dat_sum, id = "subject_int", end = "end",
  increasing = FALSE, col = "grey50", alpha = 0.2, width = 0.8
) +
  swimplot::swimmer_points(
    df = dat_swim, id = "subject_int", time = "period",
    name_col = "Pill_type", name_shape = "Pill_type", size = 4
  ) +
  ggplot2::scale_colour_manual(
    values = c("#888888", "#1b9e77", "#d95f02", "#7570b3")
  ) +
  ggplot2::labs(
    x = "Treatment period",
    y = "Subject",
    title = "Fecal fat crossover: Treatment sequence by subject"
  )
```

### Model estimates

```{r parameter-estimation}
fit <- stats::lm(formula_main, data = dat)
summary(fit)
```

### Model diagnostics

```{r adequacy-checks, fig.height=4, fig.width=6}
fit_data <- data.frame(
  fitted = stats::fitted(fit),
  residuals = stats::residuals(fit),
  std_residuals = stats::rstandard(fit)
)

ggplot2::ggplot(fit_data, ggplot2::aes(x = fitted, y = residuals)) +
  ggplot2::geom_point() +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  ggplot2::geom_smooth(se = FALSE, color = "blue") +
  ggplot2::labs(
    title = "Residuals vs Fitted",
    x = "Fitted values",
    y = "Residuals"
  ) +
  ggplot2::theme_minimal()

ggplot2::ggplot(fit_data, ggplot2::aes(sample = std_residuals)) +
  ggplot2::stat_qq() +
  ggplot2::stat_qq_line(color = "red") +
  ggplot2::labs(
    title = "Normal Q-Q",
    x = "Theoretical Quantiles",
    y = "Standardized residuals"
  ) +
  ggplot2::theme_minimal()
```

### Inference

```{r inference}
ci <- stats::confint(fit)
coefs <- summary(fit)$coefficients
pill_rows <- grep("^pilltype", rownames(coefs))
data.frame(
  term = rownames(coefs)[pill_rows],
  estimate = coefs[pill_rows, "Estimate"],
  conf_low = ci[pill_rows, 1],
  conf_high = ci[pill_rows, 2],
  p_value = coefs[pill_rows, "Pr(>|t|)"]
)
```

## Discussion

After blocking on subject identity to remove between-subject variability,
differences in mean fecal fat across pill types reflect the within-subject treatment effect
of each formulation (RMB2e Ch. 7).
This design is particularly powerful because it eliminates the dominant source of variation
(subject-to-subject differences in baseline fat excretion),
allowing a relatively small crossover study to detect meaningful differences.
The residual plots confirm the adequacy of the linear model for inference.

## Source

- UCSF Regression Methods companion data: https://regression.ucsf.edu/sites/g/files/tkssra16191/files/wysiwyg/home/data/fecfat.dta
- Book: Vittinghoff E, Glidden DV, Shiboski SC, McCulloch CE (2012). *Regression Methods in Biostatistics* (2nd edition).
