---
title: "Whickham data"
format: html
vignette: >
  %\VignetteIndexEntry{Whickham data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmb)
```

This article analyzes aggregate Whickham cohort data to illustrate how age confounding can reverse the apparent association between smoking and 20-year mortality—a classic example of Simpson's paradox described in RMB2e Chapter 3.

## Introduction

The Whickham study was a population-based prospective cohort
conducted in a town in northeast England.
Women aged 18–64 were interviewed in 1972–1974
and their vital status was ascertained 20 years later.
The crude comparison of mortality rates between smokers and non-smokers
shows the counterintuitive result that smokers appear to have lower mortality—
a finding explained by the younger age distribution of smokers at baseline.
Age is a strong confounder
because it is associated with both smoking prevalence and 20-year mortality risk (RMB2e Ch. 3).

```{r study-design}
data(rmb_datasets, package = "rmb")
rmb_datasets$study_design[rmb_datasets$object == "whickham"]
```

Is current smoking associated with 20-year all-cause mortality after adjustment for age at baseline?

### Causal assumptions

```{mermaid}
flowchart LR
  AGE[Age at baseline] --> SMOKE[Current smoking]
  AGE --> DEATH[20-year mortality]
  SMOKE --> DEATH
```

## Methods

### Study sample

```{r load-data}
data(whickham, package = "rmb")
dat <- whickham
dim(dat)
summary(dat)
```

### Statistical analysis

Because the dataset comprises 12 aggregate rows (age-group × smoking × vital-status counts),
the analysis computes stratum-specific mortality rates and odds ratios by age group,
then contrasts crude and age-adjusted estimates
to illustrate Simpson's paradox as in RMB2e Chapter 3.

```{r model-specification}
formula_crude <- vstatus ~ smoker
formula_adjusted <- vstatus ~ smoker + agegrp
list(crude = formula_crude, age_adjusted = formula_adjusted)
```

## Results

### Descriptive statistics

```{r exploratory-summaries}
dat
with(dat, tapply(nn, list(agegrp, smoker), sum))
```

```{r exploratory-plot, fig.height=4, fig.width=6}
mort_by_age <- aggregate(nn ~ agegrp + smoker + vstatus, data = dat, sum)
dead <- subset(mort_by_age, vstatus == 1)
total <- aggregate(nn ~ agegrp + smoker, data = dat, sum)
merged <- merge(dead, total, by = c("agegrp", "smoker"), suffixes = c("_dead", "_total"))
merged$mort_rate <- merged$nn_dead / merged$nn_total

barplot(
  tapply(merged$mort_rate, list(merged$agegrp, merged$smoker), mean),
  beside = TRUE,
  col = c("#1b9e77", "#d95f02"),
  legend.text = c("Non-smoker", "Smoker"),
  xlab = "Age group",
  ylab = "20-year mortality rate",
  main = "Mortality rate by age group and smoking status"
)
```

### Model estimates

```{r parameter-estimation}
fit_crude <- stats::glm(
  formula_crude,
  data = dat,
  weights = nn,
  family = stats::binomial()
)
fit_adj <- stats::glm(
  formula_adjusted,
  data = dat,
  weights = nn,
  family = stats::binomial()
)
summary(fit_crude)
summary(fit_adj)
```

### Model diagnostics

```{r adequacy-checks}
data.frame(
  model = c("crude", "age-adjusted"),
  AIC = c(stats::AIC(fit_crude), stats::AIC(fit_adj))
)
```

### Inference

```{r inference}
or_crude <- exp(stats::coef(fit_crude))
or_adj <- exp(stats::coef(fit_adj))

ci_crude <- exp(stats::confint(fit_crude))
ci_adj <- exp(stats::confint(fit_adj))

data.frame(
  model = c("crude smoker OR", "age-adjusted smoker OR"),
  odds_ratio = c(unname(or_crude["smoker"]), unname(or_adj["smoker"])),
  conf_low = c(ci_crude["smoker", 1], ci_adj["smoker", 1]),
  conf_high = c(ci_crude["smoker", 2], ci_adj["smoker", 2])
)
```

## Discussion

The crude analysis shows that smokers have lower or similar mortality to non-smokers,
illustrating Simpson's paradox:
smokers are younger on average at baseline,
so the negative confounding from age suppresses the harmful effect of smoking.
After stratification and age adjustment,
the association between smoking and mortality is attenuated or reversed within age groups,
consistent with the age-adjusted analysis reported in RMB2e Chapter 3.
This example motivates the importance of adjusting for confounders like age
before drawing causal conclusions from observational data.

## Source

- UCSF Regression Methods companion data: https://regression.ucsf.edu/sites/g/files/tkssra16191/files/wysiwyg/home/data/whickam.dta
- Book: Vittinghoff E, Glidden DV, Shiboski SC, McCulloch CE (2012). *Regression Methods in Biostatistics* (2nd edition).
