---
title: "Figure 4.1 data"
format: html
vignette: >
  %\VignetteIndexEntry{Figure 4.1 data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmb)
```

This vignette demonstrates an RMB2e-style modeling workflow for `figure4_1`, with explicit demonstration of exploratory analysis, model specification, parameter estimation, adequacy checks, and scientific interpretation.

## Study design

```{r study-design}
data(rmb_datasets, package = "rmb")
rmb_datasets$study_design[rmb_datasets$object == "figure4_1"]
```


```{r load-figure4_1}
data(figure4_1, package = "rmb")
dat <- figure4_1
dim(dat)
head(dat)
summary(dat)
```

## Variable labels

```{r labels}
var_labels <- vapply(dat, function(x) {
  lbl <- attr(x, "label")
  if (is.null(lbl) || length(lbl) == 0 || all(lbl == "")) NA_character_ else paste(as.character(lbl), collapse = "; ")
}, character(1))
stats::setNames(var_labels, names(dat))
```

## RMB2e analysis workflow

### Research question

For `figure4_1` we ask: what is the association between a primary exposure and an outcome,
after accounting for a plausible confounder?

```{r identify-variables}
# Variable-type helper functions are provided by the rmb package:
# `is_binary()` and `is_count()`.

candidate_vars <- names(dat)[vapply(dat, function(x) {
  is.numeric(x) || is.factor(x) || is.logical(x)
}, logical(1))]

continuous_vars <- candidate_vars[vapply(dat[candidate_vars], function(x) {
  is.numeric(x) && length(unique(stats::na.omit(x))) > 10
}, logical(1))]
binary_vars <- candidate_vars[vapply(dat[candidate_vars], is_binary, logical(1))]
count_vars <- candidate_vars[vapply(dat[candidate_vars], is_count, logical(1))]

outcome_var <- if (length(binary_vars) > 0) {
  binary_vars[1]
} else if (length(continuous_vars) > 0) {
  continuous_vars[1]
} else if (length(count_vars) > 0) {
  count_vars[1]
} else if (length(candidate_vars) > 0) {
  candidate_vars[1]
} else {
  NA_character_
}

remaining_vars <- setdiff(candidate_vars, outcome_var)
exposure_var <- if (length(remaining_vars) > 0) remaining_vars[1] else NA_character_
confounder_var <- if (length(remaining_vars) > 1) remaining_vars[2] else NA_character_

model_family <- if (!is.na(outcome_var) && outcome_var %in% binary_vars) {
  "binomial"
} else if (!is.na(outcome_var) && outcome_var %in% count_vars) {
  "poisson"
} else {
  "gaussian"
}

can_fit_model <- !is.na(outcome_var) && !is.na(exposure_var)

c(
  outcome = outcome_var,
  exposure = exposure_var,
  confounder = confounder_var,
  family = model_family,
  can_fit_model = can_fit_model
)
```

We use `r outcome_var` as the outcome and `r exposure_var` as the primary exposure,
with `r confounder_var` as a candidate confounder.

The goal of this worked example is to show how the chosen model family can answer the research question by quantifying the direction and strength of the exposure-outcome association after confounder adjustment.

### Causal assumptions (conceptual DAG)

```{r dag, results='asis'}
if (!is.na(exposure_var) && !is.na(confounder_var)) {
  cat(
    "```{mermaid}
",
    "flowchart LR
",
    "  C[", confounder_var, "] --> E[", exposure_var, "]
",
    "  C --> Y[", outcome_var, "]
",
    "  E --> Y
",
    "```
",
    sep = ""
  )
} else if (!is.na(exposure_var)) {
  cat(
    "```{mermaid}
",
    "flowchart LR
",
    "  E[", exposure_var, "] --> Y[", outcome_var, "]
",
    "```
",
    sep = ""
  )
} else {
  cat("Not enough variables to draw a DAG for this dataset.\n")
}
```

### Step 0: Exploratory data analysis

We start by checking the analysis sample and the marginal distributions of the selected
variables. This step demonstrates how we verify data support for modeling assumptions
before fitting regression models.

```{r exploratory-summaries}
analysis_vars <- unique(stats::na.omit(c(outcome_var, exposure_var, confounder_var)))
if (length(analysis_vars) == 0) {
  analysis_df <- data.frame()
  c(n_total = nrow(dat), n_complete = 0, p_complete = 0)
  cat("No numeric, logical, or factor variables available for analysis.\n")
} else {
  analysis_df <- dat[, analysis_vars, drop = FALSE]
  analysis_df <- stats::na.omit(analysis_df)
  c(
    n_total = nrow(dat),
    n_complete = nrow(analysis_df),
    p_complete = nrow(analysis_df) / nrow(dat)
  )
  summary(analysis_df)
}
```

```{r exploratory-plot, fig.height=4, fig.width=6}
if (nrow(analysis_df) == 0 || is.na(exposure_var) || is.na(outcome_var)) {
  plot.new()
  text(0.5, 0.5, "No EDA plot: insufficient variables or complete cases")
} else if (is.numeric(analysis_df[[outcome_var]]) &&
    is.numeric(analysis_df[[exposure_var]])) {
  plot(
    analysis_df[[exposure_var]],
    analysis_df[[outcome_var]],
    xlab = exposure_var,
    ylab = outcome_var,
    main = "Outcome vs exposure"
  )
} else if (is.numeric(analysis_df[[outcome_var]])) {
  boxplot(
    analysis_df[[outcome_var]] ~ as.factor(analysis_df[[exposure_var]]),
    xlab = exposure_var,
    ylab = outcome_var,
    main = "Outcome by exposure"
  )
} else {
  plot.new()
  text(0.5, 0.5, "No simple EDA plot available for selected variables")
}
```

### Step 1: Model specification

Following the GLM model-fitting process, we specify an unadjusted model and a model
adjusted for the candidate confounder.

```{r model-specification}
if (can_fit_model && nrow(analysis_df) > 0) {
  if (inherits(analysis_df[[outcome_var]], "haven_labelled")) {
    analysis_df[[outcome_var]] <- as.numeric(analysis_df[[outcome_var]])
  }
  if (inherits(analysis_df[[exposure_var]], "haven_labelled")) {
    analysis_df[[exposure_var]] <- as.numeric(analysis_df[[exposure_var]])
  }
  if (!is.na(confounder_var) &&
      inherits(analysis_df[[confounder_var]], "haven_labelled")) {
    analysis_df[[confounder_var]] <- as.numeric(analysis_df[[confounder_var]])
  }

  if (is.logical(analysis_df[[outcome_var]])) {
    analysis_df[[outcome_var]] <- as.integer(analysis_df[[outcome_var]])
  }
  if (is.logical(analysis_df[[exposure_var]])) {
    analysis_df[[exposure_var]] <- as.integer(analysis_df[[exposure_var]])
  }
  if (!is.na(confounder_var) && is.logical(analysis_df[[confounder_var]])) {
    analysis_df[[confounder_var]] <- as.integer(analysis_df[[confounder_var]])
  }

  if (model_family == "binomial") {
    outcome_values <- sort(unique(stats::na.omit(analysis_df[[outcome_var]])))
    if (length(outcome_values) == 2) {
      analysis_df[[outcome_var]] <- as.integer(
        analysis_df[[outcome_var]] == outcome_values[2]
      )
    }
  }

  formula_unadjusted <- stats::as.formula(paste(outcome_var, "~", exposure_var))
  formula_adjusted <- if (!is.na(confounder_var)) {
    stats::as.formula(paste(outcome_var, "~", exposure_var, "+", confounder_var))
  } else {
    formula_unadjusted
  }

  list(
    family = model_family,
    unadjusted = formula_unadjusted,
    adjusted = formula_adjusted
  )
} else {
  formula_unadjusted <- NULL
  formula_adjusted <- NULL
  cat("Model specification skipped: insufficient variables or complete cases.\n")
}
```

### Step 2: Parameter estimation

Here we estimate unadjusted and adjusted models to show how confounder adjustment can change effect estimates and uncertainty.

```{r parameter-estimation}
fit_unadjusted <- NULL
fit_adjusted <- NULL
n_coef_display <- 6

if (!is.null(formula_unadjusted) && nrow(analysis_df) >= 20) {
  fit_function <- if (model_family == "gaussian") stats::lm else stats::glm
  fit_args <- if (model_family == "gaussian") {
    list(data = analysis_df)
  } else {
    list(data = analysis_df, family = get(model_family, asNamespace("stats"))())
  }

  fit_unadjusted <- tryCatch(
    do.call(fit_function, c(list(formula = formula_unadjusted), fit_args)),
    error = identity
  )
  fit_adjusted <- tryCatch(
    do.call(fit_function, c(list(formula = formula_adjusted), fit_args)),
    error = identity
  )

  if (inherits(fit_unadjusted, "error") || inherits(fit_adjusted, "error")) {
    cat("Model fitting did not converge for this variable set.\n")
  } else {
    coef_unadjusted <- stats::coef(summary(fit_unadjusted))
    coef_adjusted <- stats::coef(summary(fit_adjusted))

    print(coef_unadjusted[seq_len(min(n_coef_display, nrow(coef_unadjusted))), , drop = FALSE])
    print(coef_adjusted[seq_len(min(n_coef_display, nrow(coef_adjusted))), , drop = FALSE])
  }
} else {
  cat("Parameter estimation skipped: fewer than 20 complete observations.\n")
}
```

### Step 3: Model adequacy checks

We examine simple fit diagnostics to assess whether the selected model is adequate for inference in this dataset.

```{r adequacy-checks}
if (!is.null(fit_adjusted) && !inherits(fit_adjusted, "error")) {
  model_comparison <- if (is.null(fit_unadjusted) || inherits(fit_unadjusted, "error") ||
      identical(formula_unadjusted, formula_adjusted)) {
    data.frame(model = "adjusted", AIC = stats::AIC(fit_adjusted))
  } else {
    data.frame(
      model = c("unadjusted", "adjusted"),
      AIC = c(stats::AIC(fit_unadjusted), stats::AIC(fit_adjusted))
    )
  }

  model_comparison

  plot(stats::fitted(fit_adjusted), stats::residuals(fit_adjusted),
    xlab = "Fitted values",
    ylab = "Residuals",
    main = "Residuals vs fitted"
  )
  abline(h = 0, lty = 2)
} else {
  cat("Model adequacy checks skipped because no fitted adjusted model is available.\n")
}
```

### Step 4: Inference

We interpret the adjusted exposure estimate directly in relation to the research question, focusing on direction, magnitude, and statistical uncertainty.

```{r inference}
if (!is.null(fit_adjusted) && !inherits(fit_adjusted, "error")) {
  coef_table <- summary(fit_adjusted)$coefficients

  exposure_rows <- grep(paste0("^", exposure_var), rownames(coef_table), value = TRUE)
  p_col <- intersect(c("Pr(>|t|)", "Pr(>|z|)"), colnames(coef_table))

  if (length(exposure_rows) > 0 && length(p_col) > 0) {
    target_row <- exposure_rows[1]
    estimate <- unname(coef_table[target_row, "Estimate"])
    p_value <- unname(coef_table[target_row, p_col[1]])

    direction <- if (is.na(estimate) || abs(estimate) < 1e-10) {
      "little"
    } else if (estimate > 0) {
      "a positive"
    } else {
      "a negative"
    }

    cat(
      "Estimated exposure effect:", round(estimate, 3),
      "(p =", signif(p_value, 3), ")\n"
    )
    cat(
      "This suggests", direction, "association between", exposure_var,
      "and", outcome_var, "in the adjusted model.\n"
    )
  } else {
    cat("Inference summary unavailable for the selected exposure term.\n")
  }
} else {
  cat("Inference skipped because no fitted adjusted model is available.\n")
}
```

### Scientific conclusions

```{r conclusions, results='asis'}
if (can_fit_model && !is.na(exposure_var) && !is.na(outcome_var) &&
    !is.null(fit_adjusted) && !inherits(fit_adjusted, "error")) {
  coef_table <- summary(fit_adjusted)$coefficients
  exposure_rows <- grep(paste0("^", exposure_var), rownames(coef_table), value = TRUE)
  p_col <- intersect(c("Pr(>|t|)", "Pr(>|z|)"), colnames(coef_table))

  if (length(exposure_rows) > 0 && length(p_col) > 0) {
    target_row <- exposure_rows[1]
    estimate <- unname(coef_table[target_row, "Estimate"])
    p_value <- unname(coef_table[target_row, p_col[1]])

    direction <- if (is.na(estimate) || abs(estimate) < 1e-10) {
      "little to no"
    } else if (estimate > 0) {
      "a positive"
    } else {
      "a negative"
    }

    evidence <- if (is.na(p_value)) {
      "with uncertain statistical evidence"
    } else if (p_value < 0.05) {
      "with statistical evidence against the null"
    } else {
      "with limited statistical evidence against the null"
    }

    cat(
      "Answer to the research question: in this dataset, the adjusted ", model_family,
      " model estimates ", direction, " association between ", exposure_var,
      " and ", outcome_var, " after accounting for ", confounder_var,
      ", ", evidence, ". These conclusions should be interpreted in light of study design, possible residual confounding, and measurement limitations.
",
      sep = ""
    )
  } else {
    cat("The fitted model does not provide a directly interpretable adjusted exposure coefficient for the selected research question.
")
  }
} else {
  cat("Scientific conclusions are limited because the dataset does not contain enough complete variables for the selected workflow.
")
}
```



## Source

- UCSF Regression Methods companion data: https://regression.ucsf.edu/sites/g/files/tkssra16191/files/wysiwyg/home/data/figure4_1.dta
- Book: Vittinghoff E, Glidden DV, Shiboski SC, McCulloch CE (2012). *Regression Methods in Biostatistics* (2nd edition).
