---
title: "PBC data"
format: html
vignette: >
  %\VignetteIndexEntry{PBC data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmb)
```

This article fits a Cox proportional hazards model to the primary biliary cirrhosis (PBC) trial data, identifying clinical predictors of mortality as described in RMB2e Chapter 6.

## Introduction

Primary biliary cirrhosis (PBC) is a chronic autoimmune cholestatic liver disease that,
if untreated, progresses to cirrhosis and liver failure.
Dickson and colleagues (1989) analyzed data from a Mayo Clinic placebo-controlled trial
of D-penicillamine in 312 PBC patients,
establishing serum bilirubin, albumin, and prothrombin time as the dominant predictors of survival.
Log-transforming these biomarkers improves linearity of the log-hazard relationship
and is standard practice in PBC prognostic modeling (RMB2e Ch. 6).

```{r study-design}
data(rmb_datasets, package = "rmb")
rmb_datasets$study_design[rmb_datasets$object == "pbc"]
```

Is higher log-bilirubin associated with shorter survival in PBC
after adjustment for log-albumin, log-prothrombin time, age, and edema status?

### Causal assumptions

```{mermaid}
flowchart LR
  BILI[Log bilirubin] --> SURV[Survival]
  ALBU[Log albumin] --> SURV
  PROT[Prothrombin time] --> SURV
  AGE --> SURV
  EDEMA --> SURV
  RX[Treatment] --> SURV
```

## Methods

### Study sample

```{r load-data}
data(pbc, package = "rmb")
dat <- pbc
dat$status <- as.integer(unclass(dat$status))
dim(dat)
summary(dat[c("years", "status", "rx", "logbili", "logalbu", "logprot", "age", "edema")])
```

### Statistical analysis

A Cox proportional hazards model is fitted with death (status == 1) as the event,
using log-transformed bilirubin, albumin, and prothrombin time
plus age and edema as predictors (RMB2e Ch. 6 / Dickson 1989).
The proportional hazards assumption is checked with scaled Schoenfeld residuals.

```{r model-specification}
formula_main <- survival::Surv(years, status == 1) ~ logbili + logalbu + logprot + age + edema
formula_main
```

## Results

### Descriptive statistics

```{r exploratory-summaries}
with(dat, table(status))
with(dat, table(rx, status))
summary(dat[c("logbili", "logalbu", "logprot", "age")])
```

```{r exploratory-plot, fig.height=4, fig.width=6}
km_all <- survival::survfit(survival::Surv(years, status == 1) ~ 1, data = dat)
plot(km_all, lwd = 2, xlab = "Years", ylab = "Survival probability",
     main = "PBC: Overall Kaplan-Meier survival curve")
```

### Model estimates

```{r parameter-estimation}
fit <- survival::coxph(formula_main, data = dat, ties = "breslow")
summary(fit)
```

### Model diagnostics

```{r adequacy-checks, fig.height=4, fig.width=6}
ph_test <- survival::cox.zph(fit)
ph_test
plot(ph_test[1])
```

### Inference

```{r inference}
hr <- exp(stats::coef(fit))
ci <- exp(stats::confint(fit))
data.frame(
  term = names(hr),
  hazard_ratio = unname(hr),
  conf_low = ci[, 1],
  conf_high = ci[, 2],
  p_value = summary(fit)$coefficients[, "Pr(>|z|)"]
)
```

## Discussion

Log-bilirubin, log-albumin, log-prothrombin time, age, and edema are all strongly associated
with mortality risk in PBC,
confirming the Mayo Clinic prognostic model reported by Dickson et al. (1989)
and discussed in RMB2e Chapter 6.
Bilirubin shows the largest hazard ratio,
consistent with its role as the primary index of hepatic synthetic dysfunction
in cholestatic disease.
The Schoenfeld residual plots should be inspected
to confirm that the proportional hazards assumption is tenable
for each predictor across the observation period.

## Source

- Dickson ER et al. (1989). Prognosis in primary biliary cirrhosis: model for decision making. *Hepatology*, 10(1), 1–7.
- UCSF Regression Methods companion data: https://regression.ucsf.edu/sites/g/files/tkssra16191/files/wysiwyg/home/data/pbc.dta
- Book: Vittinghoff E, Glidden DV, Shiboski SC, McCulloch CE (2012). *Regression Methods in Biostatistics* (2nd edition).
