---
title: "HERS data"
format: html
vignette: >
  %\VignetteIndexEntry{HERS data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmb)
```

This article reproduces HERS analyses from RMB2e Chapters 3 and 4.

## Study design

```{r study-design}
data(rmb_datasets, package = "rmb")
rmb_datasets$study_design[rmb_datasets$object == "hers"]
```

```{r load-hers}
data(hers, package = "rmb")
dat <- hers
dim(dat)
summary(dat[c("HT", "BMI", "LDL", "LDL1", "exercise", "glucose", "age", "nonwhite", "smoking", "drinkany", "statins")])
```

## RMB2e analysis workflow

### Research question

1. Is higher BMI associated with higher LDL after adjustment for demographic/lifestyle covariates?
2. Does the treatment effect of hormone therapy on LDL differ by baseline statin use?

RMB2e references: glucose-by-exercise comparison in HERS (p. 49), adjusted BMI-LDL model (p. 113), HT-by-statin interaction model and subgroup interpretation (pp. 120–121), interaction interpretation issues (p. 126).

### Causal assumptions (conceptual DAG)

```{mermaid}
flowchart LR
  BMI[BMI] --> LDL[LDL cholesterol]
  AGE[Age] --> BMI
  AGE --> LDL
  NW[Nonwhite ethnicity] --> BMI
  NW --> LDL
  SMK[Smoking] --> BMI
  SMK --> LDL
  DRINK[Alcohol use] --> BMI
  DRINK --> LDL
  HT[Randomized HT] --> LDL1[Year-1 LDL]
  STATIN[Baseline statin use] --> LDL1
  HT --> INT[HT × statin interaction]
  STATIN --> INT
  INT --> LDL1
```

### Step 0: Exploratory data analysis

RMB2e emphasizes checking simple group differences and distributions before multipredictor modeling (pp. 49, 113).

```{r exploratory-summaries}
nd <- subset(dat, diabetes == 0)

with(nd, tapply(glucose, exercise, summary))
stats::t.test(glucose ~ exercise, data = nd)

summary(nd$BMI)
summary(nd$LDL)
with(nd, cor(BMI, LDL, use = "complete.obs"))
```

```{r exploratory-plot, fig.height=4, fig.width=6}
plot(nd$BMI, nd$LDL,
     pch = 16, cex = 0.5, col = rgb(0, 0, 0, 0.2),
     xlab = "BMI (kg/m^2)", ylab = "LDL (mg/dL)",
     main = "HERS: LDL vs BMI (women without diabetes)")
abline(stats::lm(LDL ~ BMI, data = nd), col = "#d95f02", lwd = 2)
```

### Step 1: Model specification

Model A (RMB2e p. 113): adjusted linear model for LDL with BMI as primary predictor.

Model B (RMB2e pp. 120–121): interaction model for year-1 LDL with HT, statin use, and HT×statin.

```{r model-specification}
formula_bmi_ldl <- LDL ~ BMI + age + nonwhite + smoking + drinkany
formula_ht_statin <- LDL1 ~ HT * statins

list(
  bmi_ldl_model = formula_bmi_ldl,
  ht_statin_model = formula_ht_statin
)
```

### Step 2: Parameter estimation

```{r parameter-estimation}
fit_bmi_ldl <- stats::lm(formula_bmi_ldl, data = nd)
fit_ht_statin <- stats::lm(formula_ht_statin, data = dat)

summary(fit_bmi_ldl)
summary(fit_ht_statin)
```

### Step 3: Model adequacy checks

```{r adequacy-checks, fig.height=4, fig.width=6}
par(mfrow = c(1, 2))
plot(fit_bmi_ldl, which = 1)
plot(fit_bmi_ldl, which = 2)
par(mfrow = c(1, 1))
```

```{r adequacy-checks-interaction}
# Estimated HT effect among baseline statin users: beta_HT + beta_HT:statins
stats::coef(fit_ht_statin)["HT"] + stats::coef(fit_ht_statin)["HT:statins"]
```

### Step 4: Inference

```{r inference}
ci_bmi <- stats::confint(fit_bmi_ldl)
ci_int <- stats::confint(fit_ht_statin)

inf_bmi <- data.frame(
  term = names(stats::coef(fit_bmi_ldl)),
  estimate = stats::coef(fit_bmi_ldl),
  conf_low = ci_bmi[, 1],
  conf_high = ci_bmi[, 2],
  p_value = summary(fit_bmi_ldl)$coefficients[, "Pr(>|t|)"]
)

inf_int <- data.frame(
  term = names(stats::coef(fit_ht_statin)),
  estimate = stats::coef(fit_ht_statin),
  conf_low = ci_int[, 1],
  conf_high = ci_int[, 2],
  p_value = summary(fit_ht_statin)$coefficients[, "Pr(>|t|)"]
)

inf_bmi
inf_int
```

### Scientific conclusions

In this RMB2e HERS reproduction, higher BMI is associated with higher LDL after adjustment for age, race/ethnicity, smoking, and alcohol (p. 113). The effect of hormone therapy on LDL differs by baseline statin use (pp. 120–121), illustrating effect modification on the linear-model scale as discussed in Chapter 4 (p. 126).
