---
title: "FIT BMD data"
format: html
vignette: >
  %\VignetteIndexEntry{FIT BMD data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmb)
```

This article uses the longitudinal FIT dataset to assess whether alendronate treatment is associated with greater end-of-study femoral neck BMD, adjusting for baseline BMD and key risk factors (RMB2e Chapter 9).

## Introduction

The Fracture Intervention Trial (FIT) collected femoral neck BMD
at both baseline and end-of-study,
enabling assessment of treatment-related BMD change alongside fracture endpoints.
Bisphosphonates such as alendronate inhibit osteoclast-mediated bone resorption,
thereby increasing BMD;
however, the magnitude of BMD gain depends on baseline BMD,
fracture history, and patient characteristics.
Linear regression on end-of-study BMD with baseline BMD as a covariate (analysis of covariance)
is the appropriate approach for estimating treatment effects on a continuous outcome
in a randomized trial (RMB2e Ch. 9 / Black 1996).

```{r study-design}
data(rmb_datasets, package = "rmb")
rmb_datasets$study_design[rmb_datasets$object == "fitdata"]
```

Is alendronate treatment associated with greater end-of-study femoral neck BMD from baseline,
after adjustment for baseline BMD, age, BMI, and prior vertebral fracture?

### Causal assumptions

```{r dag, fig.height=4, fig.width=6}
set.seed(42)
dag <- ggdag::dagify(
  cobmd ~ tx + blbmd + age + bmi + blvfx,
  labels = c(
    cobmd = "End BMD",
    tx = "Alendronate",
    blbmd = "Baseline BMD",
    age = "Age",
    bmi = "BMI",
    blvfx = "Prior fx"
  ),
  exposure = "tx",
  outcome = "cobmd"
)
ggdag::ggdag(dag, use_labels = "label", text = FALSE) +
  ggdag::theme_dag_blank() +
  ggplot2::labs(title = "FIT data: Causal DAG")
```

## Methods

### Study sample

```{r load-data}
data(fitdata, package = "rmb")
dat <- fitdata
dim(dat)
summary(haven::zap_labels(dat[c("tx", "blbmd", "cobmd", "age", "bmi", "blvfx")]))
```

### Statistical analysis

A linear model regresses end-of-study femoral neck BMD (`cobmd`)
on treatment assignment (`tx`), baseline BMD (`blbmd`), age, BMI,
and prior vertebral fracture history (`blvfx`),
following the analysis of covariance approach recommended in RMB2e Chapter 9.

```{r model-specification}
formula_main <- cobmd ~ tx + blbmd + age + bmi + blvfx
formula_main
```

## Results

### Descriptive statistics

```{r exploratory-summaries}
with(dat, tapply(cobmd, tx, summary))
with(dat, tapply(blbmd, tx, summary))
with(dat, cor(cbind(cobmd, blbmd, age, bmi), use = "complete.obs"))
```

```{r exploratory-plot, fig.height=4, fig.width=6}
dat$tx_label <- factor(dat$tx, levels = c(0, 1), labels = c("Placebo", "Alendronate"))

ggplot2::ggplot(dat, ggplot2::aes(x = blbmd, y = cobmd, color = tx_label)) +
  ggplot2::geom_point(alpha = 0.6, size = 0.8) +
  ggplot2::scale_color_manual(values = c("#1b9e77", "#d95f02")) +
  ggplot2::labs(
    title = "FIT: End BMD vs baseline BMD by treatment",
    x = "Baseline femoral neck BMD (g/cm²)",
    y = "End-of-study femoral neck BMD (g/cm²)",
    color = NULL
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = c(0.15, 0.9))
```

### Model estimates

```{r parameter-estimation}
fit_model <- stats::lm(formula_main, data = dat)
summary(fit_model)
```

### Model diagnostics

```{r adequacy-checks, fig.height=4, fig.width=6}
fit_data <- data.frame(
  fitted = stats::fitted(fit_model),
  residuals = stats::residuals(fit_model),
  std_residuals = stats::rstandard(fit_model)
)

ggplot2::ggplot(fit_data, ggplot2::aes(x = fitted, y = residuals)) +
  ggplot2::geom_point() +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  ggplot2::geom_smooth(se = FALSE, color = "blue") +
  ggplot2::labs(
    title = "Residuals vs Fitted",
    x = "Fitted values",
    y = "Residuals"
  ) +
  ggplot2::theme_minimal()

ggplot2::ggplot(fit_data, ggplot2::aes(sample = std_residuals)) +
  ggplot2::stat_qq() +
  ggplot2::stat_qq_line(color = "red") +
  ggplot2::labs(
    title = "Normal Q-Q",
    x = "Theoretical Quantiles",
    y = "Standardized residuals"
  ) +
  ggplot2::theme_minimal()
```

### Inference

```{r inference}
ci <- stats::confint(fit_model)
coefs <- summary(fit_model)$coefficients
data.frame(
  term = rownames(coefs),
  estimate = coefs[, "Estimate"],
  conf_low = ci[, 1],
  conf_high = ci[, 2],
  p_value = coefs[, "Pr(>|t|)"]
)
```

## Discussion

Alendronate treatment is positively associated with end-of-study femoral neck BMD
after adjustment for baseline BMD and clinical characteristics,
consistent with the mechanism of bisphosphonate action
and the primary findings of Black et al. (1996).
Baseline BMD is the dominant predictor of end-of-study BMD,
as expected from the high autocorrelation of bone density measurements.
This analysis of covariance design (with baseline BMD as covariate)
is the recommended approach for estimating treatment effects on a continuous outcome
in a randomized trial with heterogeneous baseline values (RMB2e Ch. 9).

## Source

- Black DM et al. (1996). Randomised trial of effect of alendronate on risk of fracture in women with existing vertebral fractures. *Lancet*, 348(9041), 1535–1541.
- UCSF Regression Methods companion data: https://regression.ucsf.edu/sites/g/files/tkssra16191/files/wysiwyg/home/data/fitdata.dta
- Book: Vittinghoff E, Glidden DV, Shiboski SC, McCulloch CE (2012). *Regression Methods in Biostatistics* (2nd edition).
