---
title: "Table 11.8 HERS data"
format: html
vignette: >
  %\VignetteIndexEntry{Table 11.8 HERS data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmb)
```

This article models longitudinal systolic blood pressure trajectories in HERS participants without diabetes, examining whether HRT assignment and exercise interact to affect blood pressure over follow-up visits (RMB2e Chapter 11).

## Introduction

Hormone therapy and lifestyle behaviors such as exercise
may jointly influence cardiovascular risk factors in postmenopausal women.
The HERS longitudinal dataset provides repeated measurements of systolic blood pressure
over multiple visits,
enabling examination of within-subject trajectories
and potential treatment–lifestyle interactions.
RMB2e Chapter 11 uses this structure to illustrate repeated-measures regression
and missing-data approaches.

```{r study-design}
data(rmb_datasets, package = "rmb")
rmb_datasets$study_design[rmb_datasets$object == "hers_nodm_longitudinal"]
```

We ask: does regular exercise moderate the association between HRT assignment
and longitudinal systolic blood pressure in postmenopausal women without diabetes?

### Causal assumptions

```{mermaid}
flowchart LR
  HRT[HRT assignment] --> SBP[Systolic BP]
  EXERCISE[Exercise] --> SBP
  HRT --> INT[HRT × exercise]
  EXERCISE --> INT
  INT --> SBP
  BMI[BMI] --> SBP
  AGE[Age] --> SBP
```

## Methods

### Study sample

```{r load-data}
data(hers_nodm_longitudinal, package = "rmb")
dat <- hers_nodm_longitudinal
dim(dat)
summary(dat[c("sbp", "group", "exercise", "bmi", "age", "visit")])
```

### Statistical analysis

We fit a linear regression model for systolic blood pressure
with an HRT-by-exercise interaction term,
adjusting for BMI, age, smoking, and visit number.
The model ignores within-person correlation
but illustrates the interaction parameterization discussed in RMB2e Chapter 11.

```{r model-specification}
formula_sbp <- sbp ~ group * exercise + bmi + age + csmker + visit
formula_sbp
```

## Results

### Descriptive statistics

```{r exploratory-summaries}
with(dat, tapply(sbp, list(group, exercise), mean, na.rm = TRUE))
with(dat, tapply(sbp, visit, mean, na.rm = TRUE))
```

```{r exploratory-plot, fig.height=4, fig.width=6}
boxplot(sbp ~ interaction(group, exercise), data = dat,
        names = c("Pbo/No ex", "HRT/No ex", "Pbo/Ex", "HRT/Ex"),
        xlab = "Treatment × Exercise",
        ylab = "Systolic BP (mmHg)",
        main = "HERS: SBP by HRT and exercise",
        col = c("grey85", "#d95f02", "grey60", "#e6ab02"),
        las = 2)
```

The swimmer plot shows visit attendance for a stratified random sample
of 60 participants (30 per treatment arm),
with a dot at each completed visit.

```{r swimplot, fig.height=8, fig.width=6}
set.seed(42)
dat_full <- as.data.frame(dat)
first_obs <- dat_full[!duplicated(dat_full$pptid), c("pptid", "group")]
ids_0 <- sample(first_obs$pptid[first_obs$group == 0], 30)
ids_1 <- sample(first_obs$pptid[first_obs$group == 1], 30)
dat_sub <- dat_full[dat_full$pptid %in% c(ids_0, ids_1), ]

dat_ends <- aggregate(visit ~ pptid, data = dat_sub, FUN = max)
names(dat_ends)[2] <- "last_visit"
dat_sum <- merge(
  dat_ends,
  first_obs[first_obs$pptid %in% c(ids_0, ids_1), ],
  by = "pptid"
)
dat_sum$Treatment <- factor(
  dat_sum$group,
  levels = c(0, 1),
  labels = c("Placebo", "HRT")
)

swimplot::swimmer_plot(
  df = dat_sum, id = "pptid", end = "last_visit",
  name_fill = "Treatment", increasing = FALSE,
  col = "black", alpha = 0.75, width = 0.8
) +
  swimplot::swimmer_points(
    df = dat_sub, id = "pptid", time = "visit",
    shape = 21, size = 2, col = "black", fill = "white"
  ) +
  ggplot2::scale_fill_manual(values = c("#1b9e77", "#d95f02")) +
  ggplot2::labs(
    x = "HERS visit number",
    y = "Participant",
    title = "HERS: Observed visits by treatment group (n = 60 sample)"
  )
```

### Model estimates

```{r parameter-estimation}
fit_sbp <- stats::lm(formula_sbp, data = dat)
summary(fit_sbp)
```

### Model diagnostics

```{r adequacy-checks, fig.height=4, fig.width=6}
par(mfrow = c(1, 2))
plot(fit_sbp, which = 1)
plot(fit_sbp, which = 2)
par(mfrow = c(1, 1))
```

### Inference

```{r inference}
ci <- stats::confint(fit_sbp)
coef_tab <- data.frame(
  term = names(stats::coef(fit_sbp)),
  estimate = stats::coef(fit_sbp),
  conf_low = ci[, 1],
  conf_high = ci[, 2],
  p_value = summary(fit_sbp)$coefficients[, "Pr(>|t|)"]
)
coef_tab
```

## Discussion

The interaction between HRT assignment and exercise on systolic blood pressure
examines whether lifestyle behaviors modify the cardiovascular effects of hormone therapy.
As discussed in RMB2e Chapter 11,
the longitudinal structure of HERS requires accounting for within-person correlation;
the regression model presented here is an approximation useful for illustrating
interaction parameterization,
while a proper generalized estimating equations (GEE) or mixed-effects model
would be needed for valid inference accounting for repeated measures.

## Source

- [UCSF companion data](https://regression.ucsf.edu/sites/g/files/tkssra16191/files/wysiwyg/home/data/hers_nodm_longitudinal.dta)
- Book: Vittinghoff E, Glidden DV, Shiboski SC, McCulloch CE (2012). *Regression Methods in Biostatistics* (2nd edition).
