---
title: "Georgia birthweight data"
format: html
vignette: >
  %\VignetteIndexEntry{Georgia birthweight data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmb)
```

This article examines whether maternal age and interpregnancy interval are associated with infant birthweight in clustered Georgia birth records, illustrating the challenges of correlated outcomes in RMB2e Chapter 7.

## Introduction

Birthweight is a key neonatal health indicator
strongly predicted by gestational age, maternal weight, and parity-related factors.
When multiple births are recorded per mother,
infant outcomes within the same mother are correlated—
violating the independence assumption of ordinary linear regression.
The Georgia births dataset (gababies) provides a clustered sample of 1,000 birth records
from mothers with multiple deliveries,
enabling study of between- and within-mother predictors of birthweight (RMB2e Ch. 7).

```{r study-design}
data(rmb_datasets, package = "rmb")
rmb_datasets$study_design[rmb_datasets$object == "gababies"]
```

Are maternal age and interpregnancy interval associated with infant birthweight,
after accounting for clustering of births within mothers?

### Causal assumptions

```{r dag, fig.height=4, fig.width=6}
set.seed(42)
dag <- ggdag::dagify(
  bw ~ momage + interval + initwt,
  interval ~ momage,
  labels = c(
    bw = "Birthweight",
    momage = "Maternal age",
    interval = "Interpregnancy interval",
    initwt = "Initial maternal weight"
  ),
  exposure = "interval",
  outcome = "bw"
)
ggdag::ggdag(dag, use_labels = "label", text = FALSE) +
  ggdag::theme_dag_blank() +
  ggplot2::labs(title = "GABABabies: Causal DAG")
```

## Methods

### Study sample

```{r load-data}
data(gababies, package = "rmb")
dat <- gababies
dim(dat)
summary(haven::zap_labels(dat[c("bweight", "cinitage", "timesnc", "momage", "delwght")]))
```

### Statistical analysis

A linear regression of birthweight on centered maternal age at first birth (`cinitage`),
interpregnancy interval (`timesnc`), and current maternal age is fitted,
ignoring clustering for a baseline OLS estimate;
the discussion notes the need for mixed-effects models
to account for within-mother correlation (RMB2e Ch. 7).

```{r model-specification}
formula_main <- bweight ~ cinitage + timesnc + momage
formula_main
```

## Results

### Descriptive statistics

```{r exploratory-summaries}
with(dat, tapply(bweight, birthord, summary))
with(dat, c(n_mothers = length(unique(momid)), n_births = nrow(dat)))
summary(haven::zap_labels(dat[c("bweight", "cinitage", "timesnc")]))
```

```{r exploratory-plot, fig.height=4, fig.width=6}
boxplot(bweight ~ birthord, data = dat,
        xlab = "Birth order",
        ylab = "Birthweight (grams)",
        main = "Georgia births: Birthweight by birth order",
        col = "grey85")
```

### Model estimates

```{r parameter-estimation}
fit <- stats::lm(formula_main, data = dat)
summary(fit)
```

### Model diagnostics

```{r adequacy-checks, fig.height=4, fig.width=6}
par(mfrow = c(1, 2))
plot(fit, which = 1)
plot(fit, which = 2)
par(mfrow = c(1, 1))
```

### Inference

```{r inference}
ci <- stats::confint(fit)
coefs <- summary(fit)$coefficients
ci_sub <- ci[rownames(coefs), , drop = FALSE]
data.frame(
  term = rownames(coefs),
  estimate = coefs[, "Estimate"],
  conf_low = ci_sub[, 1],
  conf_high = ci_sub[, 2],
  p_value = coefs[, "Pr(>|t|)"]
)
```

## Discussion

The OLS estimates show the direction of associations
between maternal characteristics and birthweight,
but standard errors underestimate uncertainty
because births within the same mother are correlated (RMB2e Ch. 7).
Proper inference requires a mixed-effects model with a random intercept for mother,
which would partition the total variance into within-mother and between-mother components.
The centered age at first birth (`cinitage`) isolates the between-mother age effect,
while `timesnc` captures the interpregnancy interval's influence on fetal growth.

## Source

- UCSF Regression Methods companion data: https://regression.ucsf.edu/sites/g/files/tkssra16191/files/wysiwyg/home/data/gababies.dta
- Book: Vittinghoff E, Glidden DV, Shiboski SC, McCulloch CE (2012). *Regression Methods in Biostatistics* (2nd edition).
