---
title: "Phototherapy data"
format: html
vignette: >
  %\VignetteIndexEntry{Phototherapy data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmb)
```

This article applies a logistic regression workflow to the phototherapy dataset, examining whether phototherapy reduces the risk of exceeding the transfusion threshold in neonates with jaundice.

## Introduction

Neonatal jaundice (hyperbilirubinemia) affects the majority of newborns and,
when severe, can cause neurological injury.
Phototherapy is the first-line treatment to lower total serum bilirubin (TSB),
yet observational studies must adjust for gestational age and birth weight
because preterm and low-birthweight infants both receive phototherapy more often
and are at higher risk of complications.
These data were assembled for confounding-adjusted phototherapy analyses
described in RMB2e Chapter 9.

```{r study-design}
data(rmb_datasets, package = "rmb")
rmb_datasets$study_design[rmb_datasets$object == "phototherapy"]
```

Is phototherapy use associated with a reduced rate of exceeding the transfusion threshold
after adjusting for gestational age and birth weight?

### Causal assumptions

```{r dag, fig.height=4, fig.width=6}
set.seed(42)
dag <- ggdag::dagify(
  ot ~ gest + bwt + pt + tsb,
  pt ~ gest + bwt,
  labels = c(
    ot = "Over transfusion threshold",
    pt = "Phototherapy",
    gest = "Gestational age",
    bwt = "Birth weight",
    tsb = "Serum bilirubin"
  ),
  exposure = "pt",
  outcome = "ot"
)
ggdag::ggdag(dag, use_labels = "label", text = FALSE) +
  ggdag::theme_dag_blank() +
  ggplot2::labs(title = "Phototherapy: Causal DAG")
```

## Methods

### Study sample

```{r load-data}
data(phototherapy, package = "rmb")
dat <- phototherapy
dim(dat)
summary(haven::zap_labels(dat[c("phototherapy", "over_thresh", "qual_TSB", "birth_wt", "gest_age", "male", "age_days")]))
```

### Statistical analysis

Logistic regression is used to model the binary outcome `over_thresh`
(TSB exceeding the transfusion threshold),
with phototherapy as the primary predictor
and gestational age and birth weight as confounders (RMB2e Ch. 9).

```{r model-specification}
formula_main <- over_thresh ~ phototherapy + birth_wt + gest_age
formula_main
```

## Results

### Descriptive statistics

```{r exploratory-summaries}
with(dat, table(phototherapy, over_thresh))
with(dat, prop.table(table(phototherapy, over_thresh), margin = 1))
summary(haven::zap_labels(dat[c("birth_wt", "gest_age", "qual_TSB")]))
```

```{r exploratory-plot, fig.height=4, fig.width=6}
dat$phototherapy_plot <- factor(
  dat$phototherapy,
  levels = c(0, 1),
  labels = c("No", "Yes")
)

ggplot2::ggplot(dat, ggplot2::aes(x = phototherapy_plot, y = gest_age, fill = phototherapy_plot)) +
  ggplot2::geom_boxplot() +
  ggplot2::scale_fill_manual(values = c("#1b9e77", "#d95f02")) +
  ggplot2::labs(
    title = "Gestational age by phototherapy group",
    x = "Phototherapy",
    y = "Gestational age (weeks)"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "none")
```

### Model estimates

```{r parameter-estimation}
fit <- stats::glm(formula_main, data = dat, family = stats::binomial())
summary(fit)
```

### Model diagnostics

```{r adequacy-checks}
data.frame(
  model = "adjusted",
  AIC = stats::AIC(fit),
  deviance = fit$deviance,
  null_deviance = fit$null.deviance
)
```

### Inference

```{r inference}
or <- exp(stats::coef(fit))
ci <- exp(stats::confint(fit))
data.frame(
  term = names(or),
  odds_ratio = unname(or),
  conf_low = ci[, 1],
  conf_high = ci[, 2],
  p_value = summary(fit)$coefficients[, "Pr(>|z|)"]
)
```

## Discussion

After adjustment for gestational age and birth weight,
the estimated association between phototherapy use and exceeding the transfusion threshold
reflects the confounding-adjusted treatment effect described in RMB2e Chapter 9.
Because gestational age and birth weight are both strong predictors of the outcome
and strongly related to phototherapy receipt,
unadjusted comparisons would be severely confounded.
The logistic model framework here illustrates the core confounding-adjustment strategy
recommended throughout RMB2e.

## Source

- UCSF Regression Methods companion data: https://regression.ucsf.edu/sites/g/files/tkssra16191/files/wysiwyg/home/data/phototherapy.dta
- Book: Vittinghoff E, Glidden DV, Shiboski SC, McCulloch CE (2012). *Regression Methods in Biostatistics* (2nd edition).
