---
title: "Leukemia data"
format: html
vignette: >
  %\VignetteIndexEntry{Leukemia data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmb)
library(survival)
```

This article reproduces the leukemia remission survival analysis from Freireich et al. (1963), comparing 6-mercaptopurine and placebo using Kaplan-Meier curves and a log-rank test as introduced in RMB2e Chapter 3.

## Introduction

Acute lymphoblastic leukemia (ALL) in adults historically carried a poor prognosis,
with complete remission rates that were short-lived without maintenance therapy.
Freireich and colleagues (1963) conducted a landmark randomized crossover trial
comparing 6-mercaptopurine (6-MP) to placebo in patients achieving complete remission,
measuring time until relapse.
This small dataset (42 patients, 21 per group) is a canonical teaching example
for survival analysis
because it clearly illustrates the Kaplan-Meier estimator and the log-rank test (RMB2e Ch. 3).

```{r study-design}
data(rmb_datasets, package = "rmb")
rmb_datasets$study_design[rmb_datasets$object == "leuk"]
```

Does 6-MP treatment prolong leukemia remission compared to placebo?

### Causal assumptions

```{r dag, fig.height=4, fig.width=6}
set.seed(42)
dag <- ggdag::dagify(
  time_relapse ~ rx,
  labels = c(time_relapse = "Time to relapse", rx = "6-MP treatment"),
  exposure = "rx",
  outcome = "time_relapse"
)
ggdag::ggdag(dag, use_labels = "label", text = FALSE) +
  ggdag::theme_dag_blank() +
  ggplot2::labs(title = "Leukemia: Causal DAG")
```

## Methods

### Study sample

```{r load-data}
data(leuk, package = "rmb")
dat <- leuk
dim(dat)
summary(haven::zap_labels(dat[c("time", "cens", "group")]))
```

### Statistical analysis

Kaplan-Meier curves are estimated by treatment group,
and the log-rank test is used to assess the null hypothesis of equal survival functions.
The 6-MP vs placebo comparison was randomized,
so no covariate adjustment is needed for valid inference (RMB2e Ch. 3).

```{r model-specification}
surv_obj <- survival::Surv(dat$time, dat$cens)
km_formula <- surv_obj ~ group
km_formula
```

## Results

### Descriptive statistics

```{r exploratory-summaries}
with(dat, tapply(time, group, summary))
with(dat, tapply(cens, group, mean))
```

```{r exploratory-plot, fig.height=4, fig.width=6}
km_fit <- survival::survfit(survival::Surv(time, cens) ~ group, data = dat)

km_df <- data.frame(
  time = km_fit$time,
  surv = km_fit$surv,
  strata = rep(names(km_fit$strata), km_fit$strata)
)
km_df$group <- sub("group=", "", km_df$strata)

ggplot2::ggplot(km_df, ggplot2::aes(x = time, y = surv, color = group)) +
  ggplot2::geom_step(linewidth = 1) +
  ggplot2::scale_color_manual(
    values = c("#1b9e77", "#d95f02"),
    labels = c("Placebo", "6-MP")
  ) +
  ggplot2::labs(
    title = "Leukemia: Kaplan-Meier curves by treatment group",
    x = "Weeks in remission",
    y = "Remission-free probability",
    color = NULL
  ) +
  ggplot2::ylim(0, 1) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = c(0.85, 0.85))
```

All 42 participants can be shown in a swimmer plot,
displaying each individual's time in remission
with relapses marked by an ×.

```{r swimplot, fig.height=7, fig.width=6}
dat_swim <- as.data.frame(dat)
dat_swim$id <- seq_len(nrow(dat_swim))
dat_swim$time <- as.numeric(dat_swim$time)
dat_swim$group_int <- as.integer(unclass(dat_swim$group))
dat_swim$Treatment <- factor(
  dat_swim$group_int,
  levels = c(1, 2),
  labels = c("6-MP", "Placebo")
)
dat_relapse <- dat_swim[dat_swim$cens == 1, ]

swimplot::swimmer_plot(
  df = dat_swim, id = "id", end = "time",
  name_fill = "Treatment", increasing = FALSE,
  col = "black", alpha = 0.85, width = 0.8
) +
  swimplot::swimmer_points(
    df = dat_relapse, id = "id", time = "time",
    shape = 4, size = 3, col = "black"
  ) +
  ggplot2::scale_fill_manual(values = c("#d95f02", "#1b9e77")) +
  ggplot2::labs(
    x = "Weeks in remission",
    y = "Patient",
    title = "Leukemia: Duration of remission by treatment (× = relapse)"
  )
```

### Model estimates

```{r parameter-estimation}
km_fit
summary(km_fit)
```

### Model diagnostics

```{r adequacy-checks}
logrank_test <- survival::survdiff(survival::Surv(time, cens) ~ group, data = dat)
logrank_test
```

### Inference

```{r inference}
chi2 <- logrank_test$chisq
df <- length(logrank_test$n) - 1
p_val <- 1 - stats::pchisq(chi2, df = df)
data.frame(
  test = "log-rank",
  chi_squared = round(chi2, 3),
  df = df,
  p_value = signif(p_val, 3)
)
```

## Discussion

The Kaplan-Meier curves show a marked separation between the 6-MP and placebo groups
throughout the follow-up period,
and the log-rank test provides strong evidence against equal survival functions (RMB2e Ch. 3).
Median remission duration is substantially longer in the 6-MP group,
mirroring the original report by Freireich et al. (1963).
This analysis serves as a foundational example of how nonparametric survival methods
can compare treatment groups without distributional assumptions,
motivating the Cox regression approach introduced in later chapters.

## Source

- Freireich EJ et al. (1963). The effect of 6-mercaptopurine on the duration of steroid-induced remissions in acute leukemia. *Blood*, 21, 699–716.
- UCSF Regression Methods companion data: https://regression.ucsf.edu/sites/g/files/tkssra16191/files/wysiwyg/home/data/leuk.dta
- Book: Vittinghoff E, Glidden DV, Shiboski SC, McCulloch CE (2012). *Regression Methods in Biostatistics* (2nd edition).
