---
title: "Ille-et-Vilaine study data"
format: html
vignette: >
  %\VignetteIndexEntry{Ille-et-Vilaine study data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmb)
```

This article reproduces the Ille-et-Vilaine esophageal cancer case-control analysis from Tuyns et al. (1977), fitting logistic regression for case status on alcohol, tobacco, and age as in RMB2e Chapter 5.

## Introduction

Esophageal squamous cell carcinoma has historically been associated
with heavy alcohol and tobacco use.
Tuyns and colleagues (1977) conducted an unmatched case-control study
in the Ille-et-Vilaine department of France,
recruiting 200 cases and 775 controls
to quantify the joint effects of alcohol and tobacco consumption on cancer risk.
Because both exposures correlate with age and with each other,
confounding-adjusted logistic regression is essential for valid inference (RMB2e Ch. 5).

```{r study-design}
data(rmb_datasets, package = "rmb")
rmb_datasets$study_design[rmb_datasets$object == "esoph"]
```

Is heavy alcohol consumption associated with esophageal cancer
after adjustment for tobacco use and age group?

### Causal assumptions

```{r dag, fig.height=4, fig.width=6}
set.seed(42)
dag <- ggdag::dagify(
  cancer ~ age + tob + alc,
  alc ~ age,
  tob ~ age,
  labels = c(
    cancer = "Esophageal cancer",
    age = "Age group",
    tob = "Tobacco",
    alc = "Alcohol"
  ),
  exposure = "alc",
  outcome = "cancer"
)
ggdag::ggdag(dag, use_labels = "label", text = FALSE) +
  ggdag::theme_dag_blank() +
  ggplot2::labs(title = "Esoph: Causal DAG")
```

## Methods

### Study sample

```{r load-data}
data(esoph, package = "rmb")
dat <- esoph
dim(dat)
summary(haven::zap_labels(dat[c("case", "age", "agegp", "tob", "tobgp", "alc", "alcgp")]))
```

### Statistical analysis

Logistic regression models case status on grouped alcohol consumption (`alcgp`)
adjusted for grouped tobacco consumption (`tobgp`) and age group (`agegp`),
following the categorical-predictor approach in RMB2e Chapter 5.

```{r model-specification}
formula_main <- case ~ alcgp + tobgp + agegp
formula_main
```

## Results

### Descriptive statistics

```{r exploratory-summaries}
with(dat, table(alcgp, case))
with(dat, prop.table(table(alcgp, case), margin = 1))
with(dat, table(tobgp, case))
```

```{r exploratory-plot, fig.height=4, fig.width=6}
alc_rates <- with(dat, tapply(case, alcgp, mean))
alc_df <- data.frame(
  alcgp = names(alc_rates),
  rate = as.numeric(alc_rates)
)

ggplot2::ggplot(alc_df, ggplot2::aes(x = alcgp, y = rate)) +
  ggplot2::geom_col(fill = "#d95f02") +
  ggplot2::labs(
    title = "Crude case rate by alcohol consumption group",
    x = "Alcohol group (gm/day)",
    y = "Proportion cases"
  ) +
  ggplot2::theme_minimal()
```

### Model estimates

```{r parameter-estimation}
fit <- stats::glm(formula_main, data = dat, family = stats::binomial())
summary(fit)
```

### Model diagnostics

```{r adequacy-checks}
fit_alc_only <- stats::glm(case ~ alcgp, data = dat, family = stats::binomial())
stats::anova(fit_alc_only, fit, test = "Chisq")
```

### Inference

```{r inference}
or <- exp(stats::coef(fit))
ci <- exp(stats::confint(fit))
data.frame(
  term = names(or),
  odds_ratio = unname(or),
  conf_low = ci[, 1],
  conf_high = ci[, 2],
  p_value = summary(fit)$coefficients[, "Pr(>|z|)"]
)
```

## Discussion

Heavy alcohol consumption is strongly associated with esophageal cancer risk
after adjustment for tobacco use and age,
consistent with the dose-response pattern reported by Tuyns et al. (1977)
and discussed in RMB2e Chapter 5.
Tobacco consumption and older age are independently associated with elevated odds.
The likelihood-ratio test comparing the alcohol-only model to the full model
confirms that tobacco and age explain additional variation in case status
beyond alcohol alone.
These findings underscore the importance of mutual adjustment when exposures are correlated.

## Source

- Tuyns AJ et al. (1977). Esophageal cancer in Ille-et-Vilaine in relation to levels of alcohol and tobacco consumption. *Bull Cancer*, 64(1), 45–60.
- UCSF Regression Methods companion data: https://regression.ucsf.edu/sites/g/files/tkssra16191/files/wysiwyg/home/data/esoph.dta
- Book: Vittinghoff E, Glidden DV, Shiboski SC, McCulloch CE (2012). *Regression Methods in Biostatistics* (2nd edition).
