---
title: "MrOS data"
format: html
vignette: >
  %\VignetteIndexEntry{MrOS data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmb)
library(survival)
```

This article fits a Cox proportional hazards model to the MrOS dataset, evaluating whether lower bone mineral density quantiles predict higher event hazard in older men after adjustment for body weight.

## Introduction

The Osteoporotic Fractures in Men (MrOS) study enrolled community-dwelling men
aged 65 years and older at six US clinical sites beginning in 2000 (Orwoll et al. 2005).
Low bone mineral density (BMD) is a well-established predictor of fracture and mortality risk
in older men,
though body weight is a strong confounder
because heavier men tend to have higher BMD and independently different event rates.
This survival analysis applies the Cox proportional hazards framework
described in RMB2e Chapter 6.

```{r study-design}
data(rmb_datasets, package = "rmb")
rmb_datasets$study_design[rmb_datasets$object == "mros"]
```

Are lower bone mineral density quantiles associated with higher event hazard
after adjustment for body weight in older men?

### Causal assumptions

```{r dag, fig.height=4, fig.width=6}
set.seed(42)
dag <- ggdag::dagify(
  surv ~ bmd + weight,
  bmd ~ weight,
  labels = c(
    surv = "Event/death",
    bmd = "BMD quantile",
    weight = "Body weight"
  ),
  exposure = "bmd",
  outcome = "surv"
)
ggdag::ggdag(dag, use_labels = "label", text = FALSE) +
  ggdag::theme_dag_blank() +
  ggplot2::labs(title = "MrOS: Causal DAG")
```

## Methods

### Study sample

```{r load-data}
data(mros, package = "rmb")
dat <- mros
dim(dat)
summary(haven::zap_labels(dat[c("bmd3", "months", "years", "status", "weight")]))
```

### Statistical analysis

A Cox proportional hazards model is fitted
with `bmd3` (three quantiles of BMD, coded as factor) and body weight as predictors,
using follow-up time in years as the time scale (RMB2e Ch. 6 / Orwoll 2005).

```{r model-specification}
formula_main <- survival::Surv(years, status) ~ factor(bmd3) + weight
formula_main
```

## Results

### Descriptive statistics

```{r exploratory-summaries}
with(dat, table(bmd3, status))
with(dat, tapply(years, bmd3, summary))
with(dat, tapply(weight, bmd3, mean))
```

```{r exploratory-plot, fig.height=4, fig.width=6}
km_fit <- survival::survfit(survival::Surv(years, status) ~ bmd3, data = dat)
plot(
  km_fit,
  col = c("#1b9e77", "#d95f02", "#7570b3"),
  lwd = 2,
  xlab = "Years",
  ylab = "Event-free survival",
  main = "MrOS: Survival by BMD quantile"
)
legend("topright", legend = c("BMD Q1 (low)", "BMD Q2", "BMD Q3 (high)"),
       col = c("#1b9e77", "#d95f02", "#7570b3"), lwd = 2, bty = "n")
```

### Model estimates

```{r parameter-estimation}
fit <- survival::coxph(formula_main, data = dat, ties = "breslow")
summary(fit)
```

### Model diagnostics

```{r adequacy-checks}
ph_test <- survival::cox.zph(fit)
ph_test
```

### Inference

```{r inference}
hr <- exp(stats::coef(fit))
ci <- exp(stats::confint(fit))
data.frame(
  term = names(hr),
  hazard_ratio = unname(hr),
  conf_low = ci[, 1],
  conf_high = ci[, 2],
  p_value = summary(fit)$coefficients[, "Pr(>|z|)"]
)
```

## Discussion

Men in lower BMD quantiles have higher hazard of the composite event
after adjustment for body weight,
consistent with the dose-response relationship between bone density
and fracture/mortality risk described in RMB2e Chapter 6.
Weight is also independently predictive,
reflecting the complex relationship between adiposity and skeletal outcomes in older men.
These findings illustrate how Cox regression can simultaneously estimate
the independent contributions of an ordinal predictor (BMD quantile)
and a continuous confounder (weight) to event risk.

## Source

- Orwoll E et al. (2005). Design and baseline characteristics of the osteoporotic fractures in men (MrOS) study. *J Bone Miner Res*, 20(4), 678-687.
- UCSF Regression Methods companion data: https://regression.ucsf.edu/sites/g/files/tkssra16191/files/wysiwyg/home/data/mros.dta
- Book: Vittinghoff E, Glidden DV, Shiboski SC, McCulloch CE (2012). *Regression Methods in Biostatistics* (2nd edition).

