---
title: "Needle sharing data"
format: html
vignette: >
  %\VignetteIndexEntry{Needle sharing data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmb)
```

This article applies linear regression on log-transformed needle-sharing frequency in a longitudinal injection drug user panel, examining associations with depression, homelessness, HIV status, and drug use patterns (RMB2e Chapter 8).

## Introduction

Needle sharing among injection drug users (IDUs)
is a primary route of HIV and hepatitis C transmission.
Mental health disorders such as depression and unstable housing (homelessness)
have been proposed as structural and psychological drivers of risk behavior,
including receptive syringe sharing.
Log-transforming the highly right-skewed sharing count (`logshsyr`)
induces approximate normality required for linear regression inference (RMB2e Ch. 8).
HIV-positive status may suppress or increase sharing depending on perceived risk,
while polydrug use and injection frequency reflect overall addiction severity.

```{r study-design}
data(rmb_datasets, package = "rmb")
rmb_datasets$study_design[rmb_datasets$object == "needle_sharing"]
```

Are depression diagnosis and homelessness associated with greater needle-sharing frequency
among injection drug users?

### Causal assumptions

```{r dag, fig.height=4, fig.width=6}
set.seed(42)
dag <- ggdag::dagify(
  log_share ~ depress + homeless + hiv + poly + injectfreq,
  labels = c(
    log_share = "Log needle sharing",
    depress = "Depression",
    homeless = "Homelessness",
    hiv = "HIV status",
    poly = "Polydrug use",
    injectfreq = "Injection frequency"
  ),
  exposure = "depress",
  outcome = "log_share"
)
ggdag::ggdag(dag, use_labels = "label", text = FALSE) +
  ggdag::theme_dag_blank() +
  ggplot2::labs(title = "Needle sharing: Causal DAG")
```

## Methods

### Study sample

```{r load-data}
data(needle_sharing, package = "rmb")
dat <- needle_sharing
dim(dat)
summary(haven::zap_labels(dat[c("logshsyr", "dprsn_dx", "homeless", "hivstat", "polydrug", "sqrtnivd")]))
```

### Statistical analysis

Linear regression of log-transformed needle sharing (`logshsyr`)
on depression diagnosis, homelessness, HIV status, polydrug use,
and square-root injection frequency,
following the transformed-outcome approach in RMB2e Chapter 8.

```{r model-specification}
formula_main <- logshsyr ~ dprsn_dx + homeless + hivstat + polydrug + sqrtnivd
formula_main
```

## Results

### Descriptive statistics

```{r exploratory-summaries}
with(dat, tapply(logshsyr, dprsn_dx, summary))
with(dat, tapply(logshsyr, homeless, summary))
with(dat, c(
  pct_depressed = mean(dprsn_dx, na.rm = TRUE),
  pct_homeless = mean(homeless, na.rm = TRUE),
  pct_hiv_pos = mean(hivstat == 1, na.rm = TRUE)
))
```

```{r exploratory-plot, fig.height=4, fig.width=6}
dat$dprsn_label <- factor(dat$dprsn_dx, levels = c(0, 1), labels = c("No depression", "Depression"))

ggplot2::ggplot(dat, ggplot2::aes(x = dprsn_label, y = logshsyr, fill = dprsn_label)) +
  ggplot2::geom_boxplot() +
  ggplot2::scale_fill_manual(values = c("#1b9e77", "#d95f02")) +
  ggplot2::labs(
    title = "Log needle sharing by depression diagnosis",
    x = NULL,
    y = "Log needle sharing"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "none")
```

### Model estimates

```{r parameter-estimation}
fit <- stats::lm(formula_main, data = dat)
summary(fit)
```

### Model diagnostics

```{r adequacy-checks, fig.height=4, fig.width=6}
#| layout-ncol: 2
fit_data <- data.frame(
  fitted = stats::fitted(fit),
  residuals = stats::residuals(fit),
  std_residuals = stats::rstandard(fit)
)

ggplot2::ggplot(fit_data, ggplot2::aes(x = fitted, y = residuals)) +
  ggplot2::geom_point() +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  ggplot2::geom_smooth(se = FALSE, color = "blue") +
  ggplot2::labs(
    title = "Residuals vs Fitted",
    x = "Fitted values",
    y = "Residuals"
  ) +
  ggplot2::theme_minimal()

ggplot2::ggplot(fit_data, ggplot2::aes(sample = std_residuals)) +
  ggplot2::stat_qq() +
  ggplot2::stat_qq_line(color = "red") +
  ggplot2::labs(
    title = "Normal Q-Q",
    x = "Theoretical Quantiles",
    y = "Standardized residuals"
  ) +
  ggplot2::theme_minimal()
```

### Inference

```{r inference}
ci <- stats::confint(fit)
coefs <- summary(fit)$coefficients
data.frame(
  term = rownames(coefs),
  estimate = coefs[, "Estimate"],
  conf_low = ci[, 1],
  conf_high = ci[, 2],
  p_value = coefs[, "Pr(>|t|)"]
)
```

## Discussion

Depression diagnosis and homelessness are associated with higher log needle-sharing frequency,
suggesting that mental health and housing instability contribute to HIV transmission risk behaviors
among IDUs, consistent with RMB2e Chapter 8.
Polydrug use and injection frequency are also positively associated with sharing,
reflecting addiction severity.
HIV-positive status shows an association that merits careful interpretation,
as seropositive individuals may perceive less marginal risk
or may have different social network positions.
The log transformation of the skewed outcome is critical
for producing well-behaved residuals and valid inference.

## Source

- UCSF Regression Methods companion data: https://regression.ucsf.edu/sites/g/files/tkssra16191/files/wysiwyg/home/data/needle_sharing.dta
- Book: Vittinghoff E, Glidden DV, Shiboski SC, McCulloch CE (2012). *Regression Methods in Biostatistics* (2nd edition).
