---
title: "Table 11.1/11.2 HERS data"
format: html
vignette: >
  %\VignetteIndexEntry{Table 11.1/11.2 HERS data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmb)
```

This article examines missing glucose data patterns at Visit 4 in HERS participants without diabetes, illustrating missing data mechanisms described in RMB2e Chapter 11.

## Introduction

Missing data are a pervasive challenge in longitudinal clinical studies.
In HERS, fasting glucose was measured repeatedly but subject to informative missingness—
women who missed visits or had missing glucose values may differ systematically from completers.
RMB2e Chapter 11 uses this dataset to illustrate
missing-at-random (MAR) and missing-not-at-random (MNAR) mechanisms.

```{r study-design}
data(rmb_datasets, package = "rmb")
rmb_datasets$study_design[rmb_datasets$object == "hers_nodm_visit4_only"]
```

We ask: among HERS participants without diabetes at Visit 4,
is missingness of fasting glucose associated with BMI category and other baseline characteristics,
suggesting non-random missingness?

### Causal assumptions

```{r dag, fig.height=4, fig.width=6}
set.seed(42)
dag <- ggdag::dagify(
  miss ~ bmi + age + smoke + htn + drink,
  labels = c(
    miss = "Missing glucose",
    bmi = "BMI category",
    age = "Age",
    smoke = "Smoking",
    htn = "HTN meds",
    drink = "Alcohol use"
  ),
  outcome = "miss"
)
ggdag::ggdag(dag, use_labels = "label", text = FALSE) +
  ggdag::theme_dag_blank() +
  ggplot2::labs(title = "HERS visit 4: Missing data DAG")
```

## Methods

### Study sample

```{r load-data}
data(hers_nodm_visit4_only, package = "rmb")
dat <- hers_nodm_visit4_only
dim(dat)
summary(dat[c("miss_gluc", "bmi_cat", "age", "csmker", "htnmeds", "drinkany")])
```

### Statistical analysis

We fit logistic regression to predict glucose missingness (`miss_gluc`)
from BMI category, age, smoking status, hypertension medication use, and alcohol consumption.

```{r model-specification}
formula_miss <- miss_gluc ~ bmi_cat + age + csmker + drinkany + htnmeds
formula_miss
```

## Results

### Descriptive statistics

```{r exploratory-summaries}
with(dat, table(miss_gluc, useNA = "ifany"))
with(dat, prop.table(table(miss_gluc)))
with(dat, table(bmi_cat, miss_gluc))
```

```{r exploratory-plot, fig.height=4, fig.width=6}
miss_by_bmi <- with(dat, tapply(miss_gluc, bmi_cat, mean, na.rm = TRUE))
barplot(miss_by_bmi,
        xlab = "BMI category (quintile)",
        ylab = "Proportion with missing glucose",
        main = "HERS Visit 4: Glucose missingness by BMI quintile",
        col = "grey70")
```

### Model estimates

```{r parameter-estimation}
fit_miss <- stats::glm(formula_miss, data = dat, family = stats::binomial())
summary(fit_miss)
```

### Model diagnostics

```{r adequacy-checks}
data.frame(
  model = "missingness model",
  AIC = stats::AIC(fit_miss),
  n_obs = nobs(fit_miss)
)
```

### Inference

```{r inference}
or <- exp(stats::coef(fit_miss))
ci <- exp(stats::confint(fit_miss))
inf <- data.frame(
  term = names(or),
  odds_ratio = unname(or),
  conf_low = ci[, 1],
  conf_high = ci[, 2],
  p_value = summary(fit_miss)$coefficients[, "Pr(>|z|)"]
)
inf
```

## Discussion

Missingness of fasting glucose at Visit 4 is associated with BMI category
and other participant characteristics,
indicating that the missing data mechanism is not completely at random (MCAR).
This pattern motivates the multiple imputation approach demonstrated in RMB2e Chapter 11,
where MAR-based imputation accounts for the observed predictors of missingness
to produce valid inference about the glucose outcome.

## Source

- [UCSF companion data](https://regression.ucsf.edu/sites/g/files/tkssra16191/files/wysiwyg/home/data/hers_nodm_visit4_only.dta)
- Book: Vittinghoff E, Glidden DV, Shiboski SC, McCulloch CE (2012). *Regression Methods in Biostatistics* (2nd edition).
