---
title: "Figure 4.12 data"
format: html
vignette: >
  %\VignetteIndexEntry{Figure 4.12 data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmb)
```

This article uses the `figure4_12` dataset to illustrate how identical summary statistics can mask radically different residual structures, motivating visual model checking in linear regression (RMB2e Chapter 4).

## Introduction

The famous Anscombe quartet (1973) demonstrated that four datasets
with nearly identical means, variances, correlations, and regression coefficients
can exhibit fundamentally different scatter patterns—
a stark lesson that numerical summaries alone are insufficient for model assessment.
The `figure4_12` dataset contains three outcomes (`y1`, `y2`, `y3`) on a common predictor `x`,
all yielding similar fitted coefficients but very different residual diagnostics (RMB2e Ch. 4).
Examining residual plots is the only reliable way to detect misspecification
that summary statistics would miss.

```{r study-design}
data(rmb_datasets, package = "rmb")
rmb_datasets$study_design[rmb_datasets$object == "figure4_12"]
```

Do three outcomes sharing a common predictor yield the same regression coefficients
but different residual diagnostic patterns?

### Causal assumptions

```{r dag, fig.height=4, fig.width=6}
set.seed(42)
dag <- ggdag::dagify(
  y1 ~ x,
  y2 ~ x,
  y3 ~ x,
  labels = c(
    x = "Predictor x",
    y1 = "Outcome y1",
    y2 = "Outcome y2",
    y3 = "Outcome y3"
  ),
  exposure = "x",
  outcome = "y1"
)
ggdag::ggdag(dag, use_labels = "label", text = FALSE) +
  ggdag::theme_dag_blank() +
  ggplot2::labs(title = "Figure 4.12: Causal DAG")
```

## Methods

### Study sample

```{r load-data}
data(figure4_12, package = "rmb")
dat <- figure4_12
dim(dat)
summary(dat)
```

### Statistical analysis

Separate simple linear regressions of `y1`, `y2`, and `y3` on `x` are fitted,
and their coefficients, R², and residual standard errors are compared.
Residual-vs-fitted plots are then produced
to reveal the different underlying data structures (RMB2e Ch. 4).

```{r model-specification}
formula_y1 <- y1 ~ x
formula_y2 <- y2 ~ x
formula_y3 <- y3 ~ x

list(y1 = formula_y1, y2 = formula_y2, y3 = formula_y3)
```

## Results

### Descriptive statistics

```{r exploratory-summaries}
with(dat, cor(cbind(y1, y2, y3, x)))
with(dat, sapply(list(y1 = y1, y2 = y2, y3 = y3), mean))
with(dat, sapply(list(y1 = y1, y2 = y2, y3 = y3), var))
```

```{r exploratory-plot, fig.height=4, fig.width=6}
dat_long <- data.frame(
  x = rep(dat$x, 3),
  y = c(dat$y1, dat$y2, dat$y3),
  outcome = rep(c("y1", "y2", "y3"), each = nrow(dat))
)

ggplot2::ggplot(dat_long, ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_point() +
  ggplot2::geom_smooth(method = "lm", se = FALSE, color = "#d95f02", linewidth = 1) +
  ggplot2::facet_wrap(~outcome, scales = "free_y") +
  ggplot2::labs(x = "x", y = "y") +
  ggplot2::theme_minimal()
```

### Model estimates

```{r parameter-estimation}
fit1 <- stats::lm(formula_y1, data = dat)
fit2 <- stats::lm(formula_y2, data = dat)
fit3 <- stats::lm(formula_y3, data = dat)

coef_table <- data.frame(
  outcome = c("y1", "y2", "y3"),
  intercept = c(stats::coef(fit1)[1], stats::coef(fit2)[1], stats::coef(fit3)[1]),
  slope_x = c(stats::coef(fit1)[2], stats::coef(fit2)[2], stats::coef(fit3)[2]),
  r_squared = c(summary(fit1)$r.squared, summary(fit2)$r.squared, summary(fit3)$r.squared),
  sigma = c(summary(fit1)$sigma, summary(fit2)$sigma, summary(fit3)$sigma)
)
coef_table
```

### Model diagnostics

```{r adequacy-checks, fig.height=4, fig.width=6}
fit_data_long <- data.frame(
  fitted = c(stats::fitted(fit1), stats::fitted(fit2), stats::fitted(fit3)),
  residuals = c(stats::residuals(fit1), stats::residuals(fit2), stats::residuals(fit3)),
  model = rep(c("y1", "y2", "y3"), each = nrow(dat))
)

ggplot2::ggplot(fit_data_long, ggplot2::aes(x = fitted, y = residuals)) +
  ggplot2::geom_point() +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  ggplot2::geom_smooth(se = FALSE, color = "blue") +
  ggplot2::facet_wrap(~model, scales = "free") +
  ggplot2::labs(x = "Fitted values", y = "Residuals") +
  ggplot2::theme_minimal()
```

### Inference

```{r inference}
ci1 <- stats::confint(fit1)
ci2 <- stats::confint(fit2)
ci3 <- stats::confint(fit3)

data.frame(
  outcome = c("y1", "y2", "y3"),
  slope_est = c(stats::coef(fit1)["x"], stats::coef(fit2)["x"], stats::coef(fit3)["x"]),
  conf_low = c(ci1["x", 1], ci2["x", 1], ci3["x", 1]),
  conf_high = c(ci1["x", 2], ci2["x", 2], ci3["x", 2])
)
```

## Discussion

The three outcomes yield nearly identical intercepts, slopes, R² values,
and residual standard errors,
yet their residual-vs-fitted plots reveal fundamentally different patterns:
`y1` may show approximately random scatter (suitable for linear regression),
`y2` exhibits a systematic curve (suggesting a nonlinear term is needed),
and `y3` may show the influence of an outlier (RMB2e Ch. 4).
This dataset reinforces the core message of RMB2e Chapter 4:
residual plots are an indispensable complement to numerical summaries,
and model adequacy must be verified visually before drawing scientific conclusions.

## Source

- UCSF Regression Methods companion data: figure 4.12 illustrative dataset (RMB2e Chapter 4).
- Book: Vittinghoff E, Glidden DV, Shiboski SC, McCulloch CE (2012). *Regression Methods in Biostatistics* (2nd edition).
