---
title: "WCGS data"
format: html
vignette: >
  %\VignetteIndexEntry{WCGS data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmb)
```

This article reproduces key WCGS analyses from RMB2e Chapters 2, 3, and 5, modelling coronary heart disease risk from standard cardiovascular risk factors.

## Introduction

The Western Collaborative Group Study (WCGS) enrolled approximately 3,500 employed men aged 39–59 in California in 1960–1961 and followed them for roughly 8.5 years to study coronary heart disease (CHD) incidence. WCGS was among the first prospective cohort studies to establish the Type A behavior pattern as a CHD risk factor alongside classical predictors such as age, serum cholesterol, and blood pressure. These data illustrate the logistic regression framework for binary outcomes (RMB2e Chapter 5, pp. 171–177).

```{r study-design}
data(rmb_datasets, package = "rmb")
rmb_datasets$study_design[rmb_datasets$object == "wcgs"]
```

How are established cardiovascular risk factors—age, cholesterol, systolic blood pressure, BMI, smoking, and behavior pattern—jointly associated with CHD incidence in WCGS men?

### Causal assumptions

```{mermaid}
flowchart LR
  AGE[Age] --> CHD[CHD by follow-up]
  CHOL[Cholesterol] --> CHD
  SBP[Systolic BP] --> CHD
  BMI[BMI] --> CHD
  SMOKE[Smoking] --> CHD
  BEH[Behavior pattern] --> SMOKE
  BEH --> CHD
  AGE --> SBP
  AGE --> CHOL
```

## Methods

### Study sample

```{r load-data}
data(wcgs, package = "rmb")
dat <- wcgs
dim(dat)
summary(dat[c("sbp", "weight", "chd69", "age", "chol", "bmi", "smoke", "behpat")])
```

### Statistical analysis

Following RMB2e Chapter 5 (pp. 171, 175), two logistic regression models are fitted on a cleaned sample (excluding one outlier with cholesterol ≥ 645 mg/dL): a base model with five standard risk factors, and an extended model adding categorical behavior pattern to test its independent contribution via a likelihood-ratio test.

```{r model-specification}
ana <- subset(dat, chol < 645)
ana$age_10 <- ana$age / 10
ana$chol_50 <- ana$chol / 50
ana$sbp_50 <- ana$sbp / 50
ana$bmi_10 <- ana$bmi / 10
ana$smoke <- factor(ana$smoke)
ana$behpat <- factor(ana$behpat)

formula_base <- chd69 ~ age_10 + chol_50 + sbp_50 + bmi_10 + smoke
formula_behpat <- update(formula_base, . ~ . + behpat)

list(base = formula_base, with_behpat = formula_behpat)
```

## Results

### Descriptive statistics

```{r exploratory-summaries}
summary(dat$sbp)
summary(dat$weight)
with(dat, cor(sbp, weight, use = "complete.obs"))
with(dat, table(behpat, useNA = "ifany"))
with(dat, prop.table(table(chd69)))
```

```{r exploratory-plot, fig.height=4, fig.width=6}
hist(dat$sbp, breaks = 15, col = "grey85", border = "white",
     main = "WCGS systolic blood pressure", xlab = "SBP (mmHg)")
```

### Model estimates

```{r parameter-estimation}
fit_base <- stats::glm(formula_base, data = ana, family = stats::binomial())
fit_behpat <- stats::glm(formula_behpat, data = ana, family = stats::binomial())

summary(fit_base)
summary(fit_behpat)
```

### Model diagnostics

```{r adequacy-checks}
stats::anova(fit_base, fit_behpat, test = "Chisq")

pred <- stats::fitted(fit_behpat)
cut_pred <- cut(pred, breaks = stats::quantile(pred, probs = seq(0, 1, by = 0.1)), include.lowest = TRUE)
obs <- tapply(ana$chd69, cut_pred, mean)
exp_val <- tapply(pred, cut_pred, mean)
data.frame(decile = names(obs), observed = as.numeric(obs), expected = as.numeric(exp_val))
```

### Inference

```{r inference}
or <- exp(stats::coef(fit_behpat))
ci <- exp(stats::confint(fit_behpat))
inf <- data.frame(
  term = names(or),
  odds_ratio = unname(or),
  conf_low = ci[, 1],
  conf_high = ci[, 2],
  p_value = summary(fit_behpat)$coefficients[, "Pr(>|z|)"]
)
inf
```

## Discussion

Consistent with RMB2e Chapter 5 (pp. 171, 175–177), older age, higher cholesterol, higher systolic blood pressure, higher BMI, and smoking are each associated with increased CHD odds after mutual adjustment. Behavior pattern (Type A vs B) contributes additional independent information about CHD risk, as demonstrated by the likelihood-ratio test comparing the base and extended models. The decile calibration table confirms reasonable model fit across the risk spectrum, an important diagnostic step for logistic regression (RMB2e p. 175).

## Source

- UCSF Regression Methods companion data: https://regression.ucsf.edu/sites/g/files/tkssra16191/files/wysiwyg/home/data/wcgs.dta
- Book: Vittinghoff E, Glidden DV, Shiboski SC, McCulloch CE (2012). *Regression Methods in Biostatistics* (2nd edition).
