---
title: "Table 11.3 HERS data"
format: html
vignette: >
  %\VignetteIndexEntry{Table 11.3 HERS data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmb)
```

This article analyzes glucose levels at Year-1 follow-up in HERS participants, examining how HRT assignment, exercise, and BMI relate to glycemic outcomes (RMB2e Chapter 11).

## Introduction

The Heart and Estrogen/Progestin Replacement Study (HERS) enrolled postmenopausal women
with established coronary heart disease and randomized them to hormone therapy (HRT) or placebo.
Beyond its primary cardiovascular endpoints,
HERS collected repeated metabolic measurements including fasting glucose,
enabling analysis of longitudinal glycemic trajectories.
Missing data patterns in metabolic outcomes are a key methodological concern
in longitudinal studies (RMB2e Chapter 11).

```{r study-design}
data(rmb_datasets, package = "rmb")
rmb_datasets$study_design[rmb_datasets$object == "hers_long_base_visit1_only_saved"]
```

This dataset contains baseline and Visit 1 data from the HERS longitudinal structure.
We ask: among women without diabetes at baseline,
is HRT assignment associated with Year-1 fasting glucose
after adjusting for exercise and BMI?

### Causal assumptions

```{r dag, fig.height=4, fig.width=6}
set.seed(42)
dag <- ggdag::dagify(
  gluc ~ hrt + bmi + exercise + age,
  bmi ~ age,
  labels = c(
    gluc = "Year-1 glucose",
    hrt = "HRT assignment",
    bmi = "BMI",
    exercise = "Exercise >=3x/wk",
    age = "Age"
  ),
  exposure = "hrt",
  outcome = "gluc"
)
ggdag::ggdag(dag, use_labels = "label", text = FALSE) +
  ggdag::theme_dag_blank() +
  ggplot2::labs(title = "HERS baseline: Causal DAG")
```

## Methods

### Study sample

```{r load-data}
data(hers_long_base_visit1_only_saved, package = "rmb")
dat <- haven::zap_labels(hers_long_base_visit1_only_saved)
dim(dat)
ana <- subset(dat, nvisit == 1 & diabetes == 0)
dim(ana)
summary(ana[c("glucose", "group", "bmi", "exer3", "age")])
```

### Statistical analysis

We fit a linear regression model for Year-1 fasting glucose
with HRT assignment (`group`) as the primary predictor,
adjusting for BMI, exercise frequency, age, and smoking status
among women attending Visit 1 without diabetes.

```{r model-specification}
formula_gluc <- glucose ~ group + bmi + exer3 + age + csmker
formula_gluc
```

## Results

### Descriptive statistics

```{r exploratory-summaries}
with(ana, tapply(glucose, group, summary))
with(ana, tapply(glucose, exer3, mean, na.rm = TRUE))
```

```{r exploratory-plot, fig.height=4, fig.width=6}
ana$group_label <- factor(ana$group, levels = c(0, 1), labels = c("Placebo", "HRT"))

ggplot2::ggplot(ana, ggplot2::aes(x = group_label, y = glucose, fill = group_label)) +
  ggplot2::geom_boxplot() +
  ggplot2::scale_fill_manual(values = c("grey85", "#d95f02")) +
  ggplot2::labs(
    title = "HERS: Year-1 glucose by HRT assignment",
    x = "Treatment assignment",
    y = "Year-1 fasting glucose (mg/dL)"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "none")
```

### Model estimates

```{r parameter-estimation}
fit_gluc <- stats::lm(formula_gluc, data = ana)
summary(fit_gluc)
```

### Model diagnostics

```{r adequacy-checks, fig.height=4, fig.width=6}
#| layout-ncol: 2
fit_data <- data.frame(
  fitted = stats::fitted(fit_gluc),
  residuals = stats::residuals(fit_gluc),
  std_residuals = stats::rstandard(fit_gluc)
)

ggplot2::ggplot(fit_data, ggplot2::aes(x = fitted, y = residuals)) +
  ggplot2::geom_point(alpha = 0.25, size = 0.5) +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  ggplot2::geom_smooth(se = FALSE, color = "blue") +
  ggplot2::labs(
    title = "Residuals vs Fitted",
    x = "Fitted values",
    y = "Residuals"
  ) +
  ggplot2::theme_minimal()

ggplot2::ggplot(fit_data, ggplot2::aes(sample = std_residuals)) +
  ggplot2::stat_qq() +
  ggplot2::stat_qq_line(color = "red") +
  ggplot2::labs(
    title = "Normal Q-Q",
    x = "Theoretical Quantiles",
    y = "Standardized residuals"
  ) +
  ggplot2::theme_minimal()
```

### Inference

```{r inference}
ci <- stats::confint(fit_gluc)
coef_tab <- data.frame(
  term = names(stats::coef(fit_gluc)),
  estimate = stats::coef(fit_gluc),
  conf_low = ci[, 1],
  conf_high = ci[, 2],
  p_value = summary(fit_gluc)$coefficients[, "Pr(>|t|)"]
)
coef_tab
```

## Discussion

In this HERS baseline/Visit-1 longitudinal analysis,
HRT assignment does not appear to substantially change fasting glucose
in women without diabetes after adjustment for BMI, exercise, and age.
The analysis illustrates the longitudinal data structure used in RMB2e Chapter 11
for missing-data examples,
where patterns of missingness in glucose are related to baseline characteristics.
Regular exercise is associated with lower glucose levels,
consistent with its known metabolic effects.

## Source

- [UCSF companion data](https://regression.ucsf.edu/sites/g/files/tkssra16191/files/wysiwyg/home/data/hers_long_base_visit1_only_saved.dta)
- Book: Vittinghoff E, Glidden DV, Shiboski SC, McCulloch CE (2012). *Regression Methods in Biostatistics* (2nd edition).
