---
title: "UNOS data"
format: html
vignette: >
  %\VignetteIndexEntry{UNOS data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmb)
library(survival)
```

This article fits a Cox proportional hazards model to pediatric kidney transplant recipients in the UNOS registry, comparing post-transplant mortality between living-donor and cadaveric recipients (RMB2e Chapter 6).

## Introduction

Kidney transplantation offers superior outcomes compared to long-term dialysis
for children with end-stage renal disease.
Living-donor transplants are hypothesized to have better outcomes than cadaveric transplants
due to shorter cold ischemic time, better HLA matching opportunities, and planned surgery timing.
The UNOS (United Network for Organ Sharing) registry provides observational data
on outcomes for all US transplant recipients,
enabling registry-based Cox regression
adjusted for recipient and donor characteristics (RMB2e Ch. 6).

```{r study-design}
data(rmb_datasets, package = "rmb")
rmb_datasets$study_design[rmb_datasets$object == "unos"]
```

Is living-donor transplant associated with lower post-transplant mortality
compared to cadaveric transplant in pediatric kidney recipients?

### Causal assumptions

```{r dag, fig.height=4, fig.width=6}
set.seed(42)
dag <- ggdag::dagify(
  mort ~ txtype + age + agedon + cold + rej,
  labels = c(
    mort = "Post-transplant mortality",
    txtype = "Living vs cadaveric",
    age = "Age",
    agedon = "Donor age",
    cold = "Cold ischemic time",
    rej = "Rejection 1yr"
  ),
  exposure = "txtype",
  outcome = "mort"
)
ggdag::ggdag(dag, use_labels = "label", text = FALSE) +
  ggdag::theme_dag_blank() +
  ggplot2::labs(title = "UNOS: Causal DAG")
```

## Methods

### Study sample

```{r load-data}
data(unos, package = "rmb")
dat <- unos
dim(dat)
summary(haven::zap_labels(dat[c("death", "fu", "txtype", "age", "age_don", "cold_isc", "trtrej1y")]))
```

### Statistical analysis

A Cox proportional hazards model is fitted with post-transplant death as the event,
using follow-up time `fu` (years) and adjusting for transplant type,
recipient age, donor age, cold ischemic time, and first-year rejection (RMB2e Ch. 6).

```{r model-specification}
formula_main <- survival::Surv(fu, death) ~ txtype + age + age_don + cold_isc + trtrej1y
formula_main
```

## Results

### Descriptive statistics

```{r exploratory-summaries}
with(dat, table(txtype, death))
with(dat, prop.table(table(txtype, death), margin = 1))
with(dat, tapply(fu, txtype, summary))
```

```{r exploratory-plot, fig.height=4, fig.width=6}
dat$txtype_plot <- factor(dat$txtype, labels = c("Cadaveric", "Living donor"))
km_fit <- survival::survfit(survival::Surv(fu, death) ~ txtype_plot, data = dat)

survminer::ggsurvplot(
  km_fit,
  data = dat,
  title = "UNOS: Post-transplant survival by donor type",
  xlab = "Years post-transplant",
  ylab = "Survival probability",
  legend.title = NULL,
  ggtheme = ggplot2::theme_minimal(),
  palette = c("#1b9e77", "#d95f02"),
  conf.int = FALSE,
  censor = TRUE
)
```

### Model estimates

```{r parameter-estimation}
fit <- survival::coxph(formula_main, data = dat, ties = "breslow")
summary(fit)
```

### Model diagnostics

```{r adequacy-checks}
ph_test <- survival::cox.zph(fit)
ph_test
```

### Inference

```{r inference}
hr <- exp(stats::coef(fit))
ci <- exp(stats::confint(fit))
data.frame(
  term = names(hr),
  hazard_ratio = unname(hr),
  conf_low = ci[, 1],
  conf_high = ci[, 2],
  p_value = summary(fit)$coefficients[, "Pr(>|z|)"]
)
```

## Discussion

Living-donor transplantation is associated with lower post-transplant mortality
compared to cadaveric transplantation after adjustment for recipient age, donor age,
cold ischemic time, and first-year rejection,
consistent with known biological and logistical advantages of living donation (RMB2e Ch. 6).
Rejection within the first year is a strong predictor of subsequent mortality,
underscoring the prognostic importance of early immune events.
Cold ischemic time reflects graft quality at implantation
and is an important modifiable predictor that differs between transplant types.

## Source

- UCSF Regression Methods companion data: https://regression.ucsf.edu/sites/g/files/tkssra16191/files/wysiwyg/home/data/unos.dta
- Book: Vittinghoff E, Glidden DV, Shiboski SC, McCulloch CE (2012). *Regression Methods in Biostatistics* (2nd edition).
