---
title: "SOF BMD data"
format: html
vignette: >
  %\VignetteIndexEntry{SOF BMD data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmb)
```

This article models longitudinal total BMD trajectories in the SOF cohort using spline basis terms for visit time, testing whether late menopause modifies the rate of BMD decline (RMB2e Chapter 7).

## Introduction

Total bone mineral density in older women declines after menopause,
but the trajectory need not be linear across follow-up years—
the rate of loss may accelerate or decelerate over time.
The SOF2 dataset extends the baseline SOF data with spline basis terms
(`visit_spl1`, `visit_spl2`) encoding the nonlinear shape of BMD change over visits,
and includes an indicator of late menopause (after age 52)
as a potential modifier of longitudinal decline (RMB2e Ch. 7).
A marginal model for repeated BMD captures the average population trajectory.

```{r study-design}
data(rmb_datasets, package = "rmb")
rmb_datasets$study_design[rmb_datasets$object == "sof2"]
```

How does total BMD change over follow-up visits,
and is the rate of decline modified by late menopause?

### Causal assumptions

```{mermaid}
flowchart LR
  BMDBASE[Baseline BMD] --> TBMD[Total BMD]
  BMI --> TBMD
  TIME[Visit/time] --> TBMD
  MENOP[Late menopause] --> TBMD
```

## Methods

### Study sample

```{r load-data}
data(sof2, package = "rmb")
dat <- sof2
dim(dat)
summary(dat[c("totbmd", "base_totbmd", "bmi", "visit_spl1", "visit_spl2", "meno_ov_52")])
```

### Statistical analysis

A marginal linear model regresses total BMD on baseline BMD, BMI,
spline basis terms for visit, and late menopause indicator,
estimating average BMD trajectories (RMB2e Ch. 7).
Standard errors from this pooled OLS model do not account for within-subject correlation,
but the model illustrates the approach before mixed-effects extensions.

```{r model-specification}
formula_main <- totbmd ~ base_totbmd + bmi + visit_spl1 + visit_spl2 + meno_ov_52
formula_main
```

## Results

### Descriptive statistics

```{r exploratory-summaries}
with(dat, table(visit))
with(dat, tapply(totbmd, visit, mean, na.rm = TRUE))
with(dat, table(meno_ov_52))
```

```{r exploratory-plot, fig.height=4, fig.width=6}
visit_means <- with(dat, tapply(totbmd, visit, mean, na.rm = TRUE))
plot(as.integer(names(visit_means)), visit_means,
     type = "b", pch = 16, col = "#1b9e77", lwd = 2,
     xlab = "Visit number",
     ylab = "Mean total BMD (g/cm\u00b2)",
     main = "SOF2: Mean total BMD by visit")
```

### Model estimates

```{r parameter-estimation}
fit <- stats::lm(formula_main, data = dat)
summary(fit)
```

### Model diagnostics

```{r adequacy-checks, fig.height=4, fig.width=6}
plot(stats::fitted(fit), stats::residuals(fit),
     pch = 16, cex = 0.4, col = rgb(0, 0, 0, 0.2),
     xlab = "Fitted total BMD", ylab = "Residuals",
     main = "Residuals vs fitted")
abline(h = 0, lty = 2, col = "#d95f02")
```

### Inference

```{r inference}
ci <- stats::confint(fit)
coefs <- summary(fit)$coefficients
data.frame(
  term = rownames(coefs),
  estimate = coefs[, "Estimate"],
  conf_low = ci[, 1],
  conf_high = ci[, 2],
  p_value = coefs[, "Pr(>|t|)"]
)
```

## Discussion

Baseline BMD is the strongest predictor of total BMD at each subsequent visit,
reflecting the high autocorrelation of bone density over time (RMB2e Ch. 7).
The spline terms for visit capture the nonlinear shape of population-average BMD change.
Women with late menopause (after age 52) tend to have higher BMD throughout follow-up,
consistent with a longer period of estrogen exposure
before the onset of accelerated bone loss.
This marginal model ignores within-subject correlation;
proper inference for longitudinal outcomes requires mixed-effects or GEE approaches
as developed more fully in RMB2e Chapter 7.

## Source

- Cummings SR et al. (1995). Risk factors for hip fracture in white women. *NEJM*, 332(12), 767–773.
- UCSF Regression Methods companion data: https://regression.ucsf.edu/sites/g/files/tkssra16191/files/wysiwyg/home/data/sof2.dta
- Book: Vittinghoff E, Glidden DV, Shiboski SC, McCulloch CE (2012). *Regression Methods in Biostatistics* (2nd edition).
