---
title: "ACTG 019 data"
format: html
vignette: >
  %\VignetteIndexEntry{ACTG 019 data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmb)
library(survival)
```

This article reproduces the ACTG 019 survival analysis workflow used in RMB2e Chapter 6, estimating the effect of zidovudine on time to AIDS or death.

## Introduction

The AIDS Clinical Trials Group protocol 019 was a placebo-controlled randomized trial of zidovudine (ZDV) in HIV-positive adults with CD4 counts between 200 and 500 cells/mm³, conducted in the late 1980s. It was among the first trials to demonstrate a survival benefit for antiretroviral therapy in this CD4 stratum. Baseline CD4 count is a strong predictor of disease progression and is included as a covariate to increase precision, even though randomization removes confounding by design (RMB2e p. 268).

```{r study-design}
data(rmb_datasets, package = "rmb")
rmb_datasets$study_design[rmb_datasets$object == "actg019"]
```

Can zidovudine treatment reduce the hazard of AIDS or death after accounting for baseline CD4 cell count?

### Causal assumptions

```{mermaid}
flowchart LR
  RX[ZDV treatment] --> T[Time to AIDS/death]
  CD4[Baseline CD4] --> T
  CD4 --> RX
```

## Methods

### Study sample

```{r load-data}
data(actg019, package = "rmb")
dat <- actg019
dim(dat)
summary(dat[c("days", "cens", "rx", "cd4")])
```

### Statistical analysis

A Cox proportional hazards model with Breslow tie-handling is used, regressing the survival outcome on treatment assignment and baseline CD4 count (RMB2e p. 267). The proportional hazards assumption is checked with scaled Schoenfeld residuals.

```{r model-specification}
cox_formula <- survival::Surv(days, cens) ~ rx + cd4
cox_formula
```

## Results

### Descriptive statistics

```{r exploratory-summaries}
with(dat, c(
  n = length(days),
  n_events = sum(cens == 1, na.rm = TRUE),
  event_rate = mean(cens == 1, na.rm = TRUE),
  median_followup_days = median(days, na.rm = TRUE)
))

with(dat, tapply(cd4, rx, summary))
with(dat, tapply(days, rx, summary))
```

```{r exploratory-plot, fig.height=4, fig.width=6}
km_fit <- survival::survfit(survival::Surv(days, cens) ~ rx, data = dat)
plot(
  km_fit,
  col = c("#1b9e77", "#d95f02"),
  lwd = 2,
  xlab = "Days since randomization",
  ylab = "AIDS/death-free survival",
  mark.time = TRUE,
  main = "ACTG 019 Kaplan-Meier curves by treatment"
)
legend("bottomleft", legend = levels(factor(dat$rx)), col = c("#1b9e77", "#d95f02"), lwd = 2, bty = "n")
```

### Model estimates

```{r parameter-estimation}
fit_cox <- survival::coxph(cox_formula, data = dat, ties = "breslow")
summary(fit_cox)
```

### Model diagnostics

```{r adequacy-checks, fig.height=4, fig.width=6}
ph_test <- survival::cox.zph(fit_cox)
ph_test
plot(ph_test)
```

### Inference

```{r inference}
hr <- exp(stats::coef(fit_cox))
ci <- exp(stats::confint(fit_cox))
out <- data.frame(
  term = names(hr),
  hazard_ratio = unname(hr),
  conf_low = ci[, 1],
  conf_high = ci[, 2],
  p_value = summary(fit_cox)$coefficients[, "Pr(>|z|)"]
)
out
```

## Discussion

ZDV assignment is associated with a lower hazard of AIDS or death, consistent with the trial's primary findings as discussed in RMB2e Chapter 6 (pp. 224, 267). Higher baseline CD4 count is protective, reflecting the well-known relationship between immune status and disease progression. The proportional hazards assumption diagnostics (Schoenfeld residuals) should be inspected to confirm that hazard ratios are constant over follow-up time, a key requirement for valid Cox model interpretation (RMB2e p. 267).

## Source

- UCSF Regression Methods companion data: https://regression.ucsf.edu/sites/g/files/tkssra16191/files/wysiwyg/home/data/actg019.dta
- Book: Vittinghoff E, Glidden DV, Shiboski SC, McCulloch CE (2012). *Regression Methods in Biostatistics* (2nd edition).
