---
title: "MIRA data"
format: html
vignette: >
  %\VignetteIndexEntry{MIRA data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmb)
```

This article applies logistic regression to the MIRA HSV-2 dataset, assessing whether having a new sexual partner is associated with incident HSV-2 infection after adjustment for age and STI history (RMB2e Chapter 5).

## Introduction

Herpes simplex virus type 2 (HSV-2) is the leading cause of genital ulcer disease worldwide
and an important HIV cofactor.
The MIRA trial evaluated diaphragm-based HIV prevention in sub-Saharan Africa;
a secondary analysis by de Bruyn et al. (2011)
examined predictors of incident HSV-2 among participants.
Acquiring a new sexual partner substantially increases exposure risk,
but age and prior STI history confound this relationship
because older women and those with STI history are more likely to have established partners
and different HSV-2 susceptibility (RMB2e Ch. 5).

```{r study-design}
data(rmb_datasets, package = "rmb")
rmb_datasets$study_design[rmb_datasets$object == "mira_hsv"]
```

Is having a new sexual partner associated with incident HSV-2 infection
after adjustment for age category and prior STI history?

### Causal assumptions

```{mermaid}
flowchart LR
  NEWPART[New partner] --> HSV2[Incident HSV-2]
  STI[STI history] --> HSV2
  AGE[Age category] --> HSV2
  AGE --> NEWPART
```

## Methods

### Study sample

```{r load-data}
data(mira_hsv, package = "rmb")
dat <- mira_hsv
dim(dat)
summary(dat[c("hsv2", "newparts", "stihx", "agecat")])
```

### Statistical analysis

Logistic regression models incident HSV-2 as a function of new partner status,
prior STI history, and age category,
following the categorical-predictor approach in RMB2e Chapter 5.

```{r model-specification}
formula_main <- hsv2 ~ newparts + stihx + agecat
formula_main
```

## Results

### Descriptive statistics

```{r exploratory-summaries}
with(dat, table(newparts, hsv2))
with(dat, prop.table(table(newparts, hsv2), margin = 1))
with(dat, table(stihx, hsv2))
```

```{r exploratory-plot, fig.height=4, fig.width=6}
hsv_rates <- with(dat, tapply(hsv2, list(agecat, newparts), mean, na.rm = TRUE))
barplot(t(hsv_rates),
        beside = TRUE,
        col = c("#1b9e77", "#d95f02"),
        xlab = "Age category",
        ylab = "HSV-2 incidence proportion",
        main = "MIRA: HSV-2 incidence by age and new partner",
        legend.text = c("No new partner", "New partner"))
```

The swimmer plot shows follow-up timelines for a stratified random sample
of 60 participants (20 per age group),
with red dots indicating months of HSV-2 diagnosis.

```{r swimplot, fig.height=8, fig.width=7}
set.seed(42)
dat_full <- as.data.frame(dat)
first_visit <- dat_full[!duplicated(dat_full$id), c("id", "agecat")]
ids_sub <- unlist(
  tapply(
    first_visit$id, first_visit$agecat,
    function(x) sample(x, min(20L, length(x)))
  )
)
dat_sub <- dat_full[dat_full$id %in% ids_sub, ]

# Summarize to one row per subject (end = last observed month)
dat_ends <- aggregate(mos ~ id, data = dat_sub, FUN = max)
names(dat_ends)[2] <- "end"
dat_sum <- merge(dat_ends, first_visit, by = "id")
dat_sum$Age_group <- factor(
  dat_sum$agecat,
  levels = c(1, 2, 3),
  labels = c("18-24", "25-29", "30+")
)
dat_events <- dat_sub[dat_sub$hsv2 == 1, ]

swimplot::swimmer_plot(
  df = dat_sum, id = "id", end = "end",
  name_fill = "Age_group", increasing = FALSE,
  col = "black", alpha = 0.75, width = 0.8
) +
  swimplot::swimmer_points(
    df = dat_events, id = "id", time = "mos",
    shape = 21, size = 3, col = "darkred", fill = "red"
  ) +
  ggplot2::scale_fill_manual(
    values = c("#1b9e77", "#d95f02", "#7570b3")
  ) +
  ggplot2::labs(
    x = "Months of follow-up",
    y = "Participant",
    title = "MIRA: Follow-up by age group (n = 60 sample); red = HSV-2 diagnosis"
  )
```

### Model estimates

```{r parameter-estimation}
fit <- stats::glm(formula_main, data = dat, family = stats::binomial())
summary(fit)
```

### Model diagnostics

```{r adequacy-checks}
fit_unadj <- stats::glm(hsv2 ~ newparts, data = dat, family = stats::binomial())
stats::anova(fit_unadj, fit, test = "Chisq")
```

### Inference

```{r inference}
or <- exp(stats::coef(fit))
ci <- exp(stats::confint(fit))
data.frame(
  term = names(or),
  odds_ratio = unname(or),
  conf_low = ci[, 1],
  conf_high = ci[, 2],
  p_value = summary(fit)$coefficients[, "Pr(>|z|)"]
)
```

## Discussion

New partner acquisition is associated with increased odds of incident HSV-2
after adjustment for age and STI history,
consistent with the transmission dynamics of a sexually transmitted pathogen
and the findings of de Bruyn et al. (2011).
Prior STI history is a marker of sexual network exposure and susceptibility,
while age reflects accumulated exposure and possible partial immunity.
This analysis illustrates how logistic regression for binary outcomes
adjusts for multiple categorical confounders simultaneously,
a core application in RMB2e Chapter 5.

## Source

- de Bruyn G et al. (2011). Herpes simplex virus type 2 acquisition in the MIRA trial. *Sex Transm Dis*, 38(3), 207–211.
- UCSF Regression Methods companion data: https://regression.ucsf.edu/sites/g/files/tkssra16191/files/wysiwyg/home/data/mira_hsv.dta
- Book: Vittinghoff E, Glidden DV, Shiboski SC, McCulloch CE (2012). *Regression Methods in Biostatistics* (2nd edition).
