---
title: "Figure 4.6 data"
format: html
vignette: >
  %\VignetteIndexEntry{Figure 4.6 data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmb)
```

This article uses the `figure4_6` dataset to illustrate residual and influence diagnostics in linear regression, as described in RMB2e Chapter 4.

## Introduction

Residual diagnostics are an essential part of the five-step RMB regression workflow:
after fitting a model,
examining residuals against fitted values, normal quantile plots, and Cook's distance
identifies violations of linearity, homoscedasticity, and normality
that can invalidate inference.
The `figure4_6` dataset provides 100 observations of a single predictor `x`
that serves as the basis for a pedagogical illustration of these diagnostic tools (RMB2e Ch. 4).

```{r study-design}
data(rmb_datasets, package = "rmb")
rmb_datasets$study_design[rmb_datasets$object == "figure4_6"]
```

Do the standard regression diagnostic plots reveal any violations of linear model assumptions
for data simulated from a linear relationship with normally distributed errors?

### Causal assumptions

```{r dag, fig.height=4, fig.width=6}
set.seed(42)
dag <- ggdag::dagify(
  y ~ x,
  labels = c(y = "Outcome y", x = "Predictor x"),
  exposure = "x",
  outcome = "y"
)
ggdag::ggdag(dag, use_labels = "label", text = FALSE) +
  ggdag::theme_dag_blank() +
  ggplot2::labs(title = "Figure 4.6: Causal DAG")
```

## Methods

### Study sample

```{r load-data}
data(figure4_6, package = "rmb")
dat <- figure4_6
dim(dat)
summary(dat)
```

### Statistical analysis

A simple linear regression of a simulated outcome `y`
(constructed from `x` with normally distributed errors) is fitted,
and the full suite of diagnostic plots from RMB2e Chapter 4 is produced:
residuals vs fitted, normal Q-Q, scale-location, and Cook's distance (RMB2e Ch. 4).

```{r model-specification}
set.seed(42)
y <- 2 + 1.5 * dat$x + stats::rnorm(100)
demo_df <- data.frame(x = dat$x, y = y)

formula_main <- y ~ x
formula_main
```

## Results

### Descriptive statistics

```{r exploratory-summaries}
summary(demo_df)
with(demo_df, cor(x, y))
```

```{r exploratory-plot, fig.height=4, fig.width=6}
ggplot2::ggplot(demo_df, ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_point(alpha = 0.4, size = 1.2) +
  ggplot2::geom_smooth(method = "lm", se = FALSE, color = "#d95f02", linewidth = 1) +
  ggplot2::labs(
    title = "Figure 4.6: Simulated data for diagnostic illustration",
    x = "x",
    y = "y (simulated)"
  ) +
  ggplot2::theme_minimal()
```

### Model estimates

```{r parameter-estimation}
fit <- stats::lm(formula_main, data = demo_df)
summary(fit)
```

### Model diagnostics

```{r adequacy-checks, fig.height=7, fig.width=8}
ggfortify_autoplot_lm <- getFromNamespace("autoplot.lm", "ggfortify")
ggfortify_autoplot_lm(
  fit,
  which = 1:4,
  ncol = 2
) +
  ggplot2::theme_minimal()
```

### Inference

```{r inference}
ci <- stats::confint(fit)
coefs <- summary(fit)$coefficients
data.frame(
  term = rownames(coefs),
  estimate = coefs[, "Estimate"],
  conf_low = ci[, 1],
  conf_high = ci[, 2],
  p_value = coefs[, "Pr(>|t|)"]
)
```

## Discussion

When data truly follow a linear model with homoscedastic normal errors,
the four standard diagnostic plots should show:
no pattern in residuals vs fitted,
points falling close to the diagonal in the Q-Q plot,
roughly constant spread in the scale-location plot,
and no observations with large Cook's distances (RMB2e Ch. 4).
This simulated dataset with correct model specification
provides a baseline reference for what "good" diagnostics look like,
against which practitioners can compare plots from real datasets
where assumptions may be violated.

## Source

- UCSF Regression Methods companion data: figure 4.6 illustrative dataset (RMB2e Chapter 4).
- Book: Vittinghoff E, Glidden DV, Shiboski SC, McCulloch CE (2012). *Regression Methods in Biostatistics* (2nd edition).
