---
title: "ACTG 019 data"
format: html
vignette: >
  %\VignetteIndexEntry{ACTG 019 data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmb)
library(survival)
```

This article reproduces the ACTG 019 survival analysis workflow used in RMB2e Chapter 6, estimating the effect of zidovudine on time to AIDS or death.

## Introduction

The AIDS Clinical Trials Group protocol 019 was a placebo-controlled randomized trial
of zidovudine (ZDV) in HIV-positive adults with CD4 counts between 200 and 500 cells/mm³,
conducted in the late 1980s.
It was among the first trials to demonstrate a survival benefit for antiretroviral therapy
in this CD4 stratum.
Baseline CD4 count is a strong predictor of disease progression
and is included as a covariate to increase precision,
even though randomization removes confounding by design (RMB2e p. 268).

```{r study-design}
data(rmb_datasets, package = "rmb")
rmb_datasets$study_design[rmb_datasets$object == "actg019"]
```

Can zidovudine treatment reduce the hazard of AIDS or death
after accounting for baseline CD4 cell count?

### Causal assumptions

:::{#fig-actg019-dag}
```{r dag, fig.height=4, fig.width=6}
set.seed(42)
dag <- ggdag::dagify(
  time_aids ~ rx + cd4,
  rx ~ cd4,
  labels = c(
    time_aids = "Time to AIDS/death",
    rx = "ZDV treatment",
    cd4 = "Baseline CD4"
  ),
  exposure = "rx",
  outcome = "time_aids"
)
ggdag::ggdag(dag, use_labels = "label", text = FALSE) +
  ggdag::theme_dag_blank() +
  ggplot2::labs(title = "ACTG 019: Causal DAG")
```
Directed acyclic graph for the ACTG 019 treatment-outcome structure.
:::

## Methods

### Study sample

```{r load-data}
data(actg019, package = "rmb")
dat <- actg019
dim(dat)
summary(haven::zap_labels(dat[c("days", "cens", "rx", "cd4")]))
```

### Statistical analysis

A Cox proportional hazards model with Breslow tie-handling is used,
regressing the survival outcome on treatment assignment and baseline CD4 count (RMB2e p. 267).
The proportional hazards assumption is checked with scaled Schoenfeld residuals.

```{r model-specification}
cox_formula <- survival::Surv(days, cens) ~ rx + cd4
cox_formula
```

## Results

### Descriptive statistics

```{r exploratory-summaries}
with(dat, c(
  n = length(days),
  n_events = sum(cens == 1, na.rm = TRUE),
  event_rate = mean(cens == 1, na.rm = TRUE),
  median_followup_days = median(days, na.rm = TRUE)
))

with(dat, tapply(cd4, rx, summary))
with(dat, tapply(days, rx, summary))
```

:::{#fig-actg019-km}
```{r exploratory-plot, fig.height=4, fig.width=6}
dat$rx_plot <- factor(dat$rx, levels = c(0, 1), labels = c("Placebo", "ZDV"))
km_fit <- survival::survfit(survival::Surv(days, cens) ~ rx_plot, data = dat)

survminer::ggsurvplot(
  km_fit,
  data = dat,
  title = "ACTG 019 Kaplan-Meier curves by treatment",
  xlab = "Days since randomization",
  ylab = "AIDS/death-free survival",
  legend.title = "Treatment",
  ggtheme = ggplot2::theme_minimal(),
  palette = c("#1b9e77", "#d95f02"),
  conf.int = FALSE,
  censor = TRUE
)
```
Kaplan-Meier survival curves by randomized treatment arm.
:::

The swimmer plot below shows individual follow-up times
for a random sample of 80 participants (40 per arm),
with × marks indicating AIDS/death events.

:::{#fig-actg019-swimmer}
```{r swimplot, fig.height=10, fig.width=7}
set.seed(42)
dat_swim <- as.data.frame(dat)
dat_swim$days <- as.numeric(dat_swim$days)
dat_swim$rx_int <- as.integer(unclass(dat_swim$rx))
dat_swim$cens_int <- as.integer(unclass(dat_swim$cens))
ids_0 <- sample(dat_swim$id[dat_swim$rx_int == 0], 40)
ids_1 <- sample(dat_swim$id[dat_swim$rx_int == 1], 40)
dat_sub <- dat_swim[dat_swim$id %in% c(ids_0, ids_1), ]
dat_sub$Treatment <- factor(
  dat_sub$rx_int,
  levels = c(0, 1),
  labels = c("Placebo", "ZDV")
)
dat_events <- dat_sub[dat_sub$cens_int == 1, ]

swimplot::swimmer_plot(
  df = dat_sub, id = "id", end = "days",
  name_fill = "Treatment", increasing = FALSE,
  col = "black", alpha = 0.75, width = 0.8
) +
  swimplot::swimmer_points(
    df = dat_events, id = "id", time = "days",
    shape = 4, size = 2, col = "black"
  ) +
  ggplot2::scale_fill_manual(values = c("#1b9e77", "#d95f02")) +
  ggplot2::labs(
    x = "Days since randomization",
    y = "Patient",
    title = "ACTG 019: Follow-up timeline by treatment (n = 80 sample; X = AIDS/death)"
  )
```
Swimmer plot of follow-up time with event markers in a balanced sample.
:::

### Model estimates

```{r parameter-estimation}
fit_cox <- survival::coxph(cox_formula, data = dat, ties = "breslow")
summary(fit_cox)
```

### Model diagnostics

:::{#fig-actg019-ph}
```{r adequacy-checks, fig.height=4, fig.width=6}
ph_test <- survival::cox.zph(fit_cox)
ph_test

survminer::ggcoxzph(
  ph_test,
  ggtheme = ggplot2::theme_minimal()
)
```
Scaled Schoenfeld residual diagnostics for proportional hazards assumptions.
:::

### Inference

:::{#tbl-actg019-inference}
```{r inference}
hr <- exp(stats::coef(fit_cox))
ci <- exp(stats::confint(fit_cox))
knitr::kable(data.frame(
  term = names(hr),
  hazard_ratio = unname(hr),
  conf_low = ci[, 1],
  conf_high = ci[, 2],
  p_value = summary(fit_cox)$coefficients[, "Pr(>|z|)"]
), digits = 3)
```
Cox model hazard-ratio estimates with 95% confidence intervals and p-values.
:::

## Discussion

ZDV assignment is associated with a lower hazard of AIDS or death,
consistent with the trial's primary findings as discussed in RMB2e Chapter 6 (pp. 224, 267).
Higher baseline CD4 count is protective,
reflecting the well-known relationship between immune status and disease progression.
The proportional hazards assumption diagnostics (Schoenfeld residuals) should be inspected
to confirm that hazard ratios are constant over follow-up time,
a key requirement for valid Cox model interpretation (RMB2e p. 267).

## Source

- UCSF Regression Methods companion data: https://regression.ucsf.edu/sites/g/files/tkssra16191/files/wysiwyg/home/data/actg019.dta
- Book: Vittinghoff E, Glidden DV, Shiboski SC, McCulloch CE (2012). *Regression Methods in Biostatistics* (2nd edition).
