---
title: "Figure 4.1 data"
format: html
vignette: >
  %\VignetteIndexEntry{Figure 4.1 data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmb)
```

This article uses the `figure4_1` dataset to illustrate polynomial regression and the importance of allowing nonlinear predictor-outcome relationships, as introduced in RMB2e Chapter 4.

## Introduction

Simple linear regression assumes that the outcome changes
at a constant rate per unit increase in the predictor.
When the true relationship is curved,
fitting a linear model yields biased estimates and poor predictions.
Polynomial regression addresses this by including higher-order terms (e.g., $x^2$)
in the linear predictor.
The `figure4_1` dataset contains only the squared predictor term `x2`,
making it a compact teaching dataset for demonstrating the quadratic model form (RMB2e Ch. 4).

```{r study-design}
data(rmb_datasets, package = "rmb")
rmb_datasets$study_design[rmb_datasets$object == "figure4_1"]
```

Can a polynomial regression model better fit a nonlinear data pattern
compared to a simple linear model?

### Causal assumptions

```{mermaid}
flowchart LR
  X[Predictor x] --> Y[Outcome y]
  X2[x squared] --> Y
```

## Methods

### Study sample

```{r load-data}
data(figure4_1, package = "rmb")
dat <- figure4_1
dim(dat)
summary(dat)
```

### Statistical analysis

Because the dataset contains only the squared predictor `x2` (n = 8),
we reconstruct `x = sqrt(x2)` and simulate a noisy outcome
to illustrate fitting both a linear and a quadratic model.
A simulated outcome with a mild negative quadratic component
demonstrates how polynomial regression compares to simple linear regression
on residual diagnostics (RMB2e Ch. 4).

```{r model-specification}
set.seed(42)
x <- sqrt(dat$x2)
y <- 2 + 1.5 * x - 0.1 * x^2 + stats::rnorm(8, 0, 0.5)
demo_df <- data.frame(x = x, x2 = dat$x2, y = y)

formula_linear <- y ~ x
formula_quad <- y ~ x + x2

list(linear = formula_linear, quadratic = formula_quad)
```

## Results

### Descriptive statistics

```{r exploratory-summaries}
summary(demo_df)
with(demo_df, cor(y, x))
with(demo_df, cor(y, x2))
```

```{r exploratory-plot, fig.height=4, fig.width=6}
plot(demo_df$x, demo_df$y,
     pch = 16, cex = 1.2, col = "#1b9e77",
     xlab = "x (sqrt of x2)",
     ylab = "y (simulated outcome)",
     main = "Figure 4.1: Illustrative polynomial regression data")
x_seq <- seq(min(demo_df$x), max(demo_df$x), length.out = 100)
```

### Model estimates

```{r parameter-estimation}
fit_lin <- stats::lm(formula_linear, data = demo_df)
fit_quad <- stats::lm(formula_quad, data = demo_df)

summary(fit_lin)
summary(fit_quad)
```

### Model diagnostics

```{r adequacy-checks, fig.height=4, fig.width=6}
par(mfrow = c(1, 2))
plot(fit_lin, which = 1, main = "Linear model residuals")
plot(fit_quad, which = 1, main = "Quadratic model residuals")
par(mfrow = c(1, 1))
```

### Inference

```{r inference}
ci_lin <- stats::confint(fit_lin)
ci_quad <- stats::confint(fit_quad)

data.frame(
  model = c("linear x", "quad x", "quad x2"),
  estimate = c(
    stats::coef(fit_lin)["x"],
    stats::coef(fit_quad)["x"],
    stats::coef(fit_quad)["x2"]
  ),
  conf_low = c(ci_lin["x", 1], ci_quad["x", 1], ci_quad["x2", 1]),
  conf_high = c(ci_lin["x", 2], ci_quad["x", 2], ci_quad["x2", 2])
)
```

## Discussion

Including the squared term `x2` changes the estimated slope for `x`
compared to the linear-only model,
illustrating how omitting a curved component can distort the linear coefficient (RMB2e Ch. 4).
The residual plots for the quadratic model show less systematic pattern
than those for the linear model, suggesting better fit.
This toy example motivates the model-checking workflow of Chapter 4:
plotting residuals against fitted values is a key diagnostic
for detecting nonlinearity that a simple linear term cannot capture.

## Source

- UCSF Regression Methods companion data: figure 4.1 illustrative dataset (RMB2e Chapter 4).
- Book: Vittinghoff E, Glidden DV, Shiboski SC, McCulloch CE (2012). *Regression Methods in Biostatistics* (2nd edition).
