---
title: "ACTG 019 data"
format: html
vignette: >
  %\VignetteIndexEntry{ACTG 019 data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmb)
library(survival)
```

This article reproduces the ACTG 019 survival analysis workflow used in RMB2e Chapter 6.

## Study design

```{r study-design}
data(rmb_datasets, package = "rmb")
rmb_datasets$study_design[rmb_datasets$object == "actg019"]
```

```{r load-actg019}
data(actg019, package = "rmb")
dat <- actg019
dim(dat)
summary(dat)
```

## RMB2e analysis workflow

### Research question

Can zidovudine treatment reduce the hazard of AIDS/death after accounting for baseline CD4 cell count?

RMB2e references: trial context and survival-data motivation (p. 224), Cox modeling details and bootstrap CI discussion (p. 267), coding sensitivity exercises (p. 278).

### Causal assumptions (conceptual DAG)

```{mermaid}
flowchart LR
  RX[ZDV treatment (rx)] --> T[Time to AIDS/death]
  CD4[Baseline CD4] --> T
  CD4 --> RX
```

Because ACTG 019 was randomized, treatment is unconfounded in expectation; CD4 is included as a strong prognostic baseline covariate (RMB2e p. 268).

### Step 0: Exploratory data analysis

RMB2e context: variable follow-up and event-time structure in ACTG 019 (p. 224).

```{r exploratory-summaries}
with(dat, c(
  n = length(days),
  n_events = sum(cens == 1, na.rm = TRUE),
  event_rate = mean(cens == 1, na.rm = TRUE),
  median_followup_days = median(days, na.rm = TRUE)
))

with(dat, tapply(cd4, rx, summary))
with(dat, tapply(days, rx, summary))
```

```{r exploratory-km, fig.height=4, fig.width=6}
km_fit <- survival::survfit(survival::Surv(days, cens) ~ rx, data = dat)
plot(
  km_fit,
  col = c("#1b9e77", "#d95f02"),
  lwd = 2,
  xlab = "Days since randomization",
  ylab = "AIDS/death-free survival",
  mark.time = TRUE,
  main = "ACTG 019 Kaplan-Meier curves by treatment"
)
legend("bottomleft", legend = levels(factor(dat$rx)), col = c("#1b9e77", "#d95f02"), lwd = 2, bty = "n")
```

### Step 1: Model specification

RMB2e model: Cox proportional hazards model with treatment (`rx`) and baseline CD4 (`cd4`) as predictors (p. 267).

```{r model-specification}
cox_formula <- survival::Surv(days, cens) ~ rx + cd4
cox_formula
```

### Step 2: Parameter estimation

```{r parameter-estimation}
fit_cox <- survival::coxph(cox_formula, data = dat, ties = "breslow")
summary(fit_cox)
```

### Step 3: Model adequacy checks

```{r adequacy-checks}
ph_test <- survival::cox.zph(fit_cox)
ph_test
```

```{r adequacy-plot, fig.height=4, fig.width=6}
plot(ph_test)
```

### Step 4: Inference

RMB2e discusses Wald and bootstrap interval interpretation for this model family (p. 267).

```{r inference}
hr <- exp(stats::coef(fit_cox))
ci <- exp(stats::confint(fit_cox))
out <- data.frame(
  term = names(hr),
  hazard_ratio = unname(hr),
  conf_low = ci[, 1],
  conf_high = ci[, 2],
  p_value = summary(fit_cox)$coefficients[, "Pr(>|z|)"]
)
out
```

### Scientific conclusions

In this RMB2e ACTG 019 reproduction, treatment assignment to ZDV is associated with a lower hazard of AIDS/death, while higher baseline CD4 is associated with lower hazard. These are the same scientific directions emphasized in Chapter 6 (pp. 224, 267).
