---
title: "NHANES data"
format: html
vignette: >
  %\VignetteIndexEntry{NHANES data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmb)
library(survey)
```

This article applies survey-weighted logistic regression to NHANES III data to estimate associations between age, sex, and race/ethnicity and diabetes prevalence in a nationally representative US sample (RMB2e Chapter 12).

## Introduction

The Third National Health and Nutrition Examination Survey (NHANES III, 1988–1994)
used a complex multistage probability sample
to produce nationally representative estimates of health and nutrition in the US population.
Because sampling probabilities differ across demographic strata,
unweighted analyses would give biased prevalence estimates and incorrect standard errors.
Survey-weighted regression using sampling weights, primary sampling units (PSUs), and strata
is required for valid inference from NHANES III data (RMB2e Ch. 12).

```{r study-design}
data(rmb_datasets, package = "rmb")
rmb_datasets$study_design[rmb_datasets$object == "nhanes3"]
```

Are age, sex, and race/ethnicity associated with diabetes prevalence
in a nationally representative US sample?

### Causal assumptions

```{r dag, fig.height=4, fig.width=6}
set.seed(42)
dag <- ggdag::dagify(
  dm ~ age + sex + aa + ma,
  labels = c(
    dm = "Diabetes",
    age = "Age per 10yr",
    sex = "Female",
    aa = "African American",
    ma = "Mexican American"
  ),
  outcome = "dm"
)
ggdag::ggdag(dag, use_labels = "label", text = FALSE) +
  ggdag::theme_dag_blank() +
  ggplot2::labs(title = "NHANES III: Causal DAG")
```

## Methods

### Study sample

```{r load-data}
data(nhanes3, package = "rmb")
dat <- nhanes3
dim(dat)
summary(dat[c("diabetes", "female", "aframer", "mexamer", "age10")])
```

### Statistical analysis

A survey design object is constructed using PSUs, strata, and sampling weights,
then survey-weighted logistic regression is fitted for diabetes
on age (per 10 years), sex, African American ethnicity, and Mexican American ethnicity
(RMB2e Ch. 12).

```{r model-specification}
nhanes_design <- survey::svydesign(
  id = ~sdppsu6,
  strata = ~sdpstra6,
  weights = ~wtpfqx6,
  data = dat,
  nest = TRUE
)
formula_main <- diabetes ~ female + aframer + mexamer + age10
formula_main
```

## Results

### Descriptive statistics

```{r exploratory-summaries}
survey::svymean(~diabetes, nhanes_design)
survey::svyby(~diabetes, ~aframer, nhanes_design, survey::svymean)
survey::svyby(~diabetes, ~mexamer, nhanes_design, survey::svymean)
```

```{r exploratory-plot, fig.height=4, fig.width=6}
dm_by_age <- survey::svyby(~diabetes, ~age10, nhanes_design, survey::svymean)
plot(dm_by_age$age10, dm_by_age$diabetes,
     type = "b", pch = 16, col = "#1b9e77", lwd = 2,
     xlab = "Age group (per 10yr)",
     ylab = "Weighted diabetes prevalence",
     main = "NHANES III: Diabetes prevalence by age (survey-weighted)")
```

### Model estimates

```{r parameter-estimation}
fit <- survey::svyglm(formula_main, design = nhanes_design, family = stats::binomial())
summary(fit)
```

### Model diagnostics

```{r adequacy-checks}
fit_age_only <- survey::svyglm(diabetes ~ age10, design = nhanes_design, family = stats::binomial())
stats::anova(fit_age_only, fit, test = "Chisq")
```

### Inference

```{r inference}
or <- exp(stats::coef(fit))
ci <- exp(stats::confint(fit))
data.frame(
  term = names(or),
  odds_ratio = unname(or),
  conf_low = ci[, 1],
  conf_high = ci[, 2],
  p_value = summary(fit)$coefficients[, "Pr(>|t|)"]
)
```

## Discussion

Older age, African American ethnicity, and Mexican American ethnicity are all associated
with higher diabetes prevalence in the US population after mutual adjustment,
consistent with well-documented patterns of diabetes disparities (RMB2e Ch. 12).
Using survey weights is essential:
the complex NHANES III sampling design over-represents certain demographic groups,
and unweighted estimates would misrepresent national prevalence.
This analysis illustrates the survey-weighted GLM approach that is required
whenever data come from complex probability samples with unequal selection probabilities.

## Source

- UCSF Regression Methods companion data: https://regression.ucsf.edu/sites/g/files/tkssra16191/files/wysiwyg/home/data/nhanes3.dta
- Book: Vittinghoff E, Glidden DV, Shiboski SC, McCulloch CE (2012). *Regression Methods in Biostatistics* (2nd edition).
